Theory of active self-organization of dense nematic structures in the actin cytoskeleton

Version 1

Abstract

The actin cytoskeleton is remarkably adaptable and multifunctional. It often organizes into nematic bundles such as contractile rings or stress fibers. However, how a uniform and isotropic actin gel self-organizes into dense nematic bundles is not fully understood. Here, using an active gel model accounting for nematic order and density variations, we identify an active patterning mechanism leading to localized dense nematic structures. Linear stability analysis and nonlinear finite element simulations establish the conditions for nematic bundle self-assembly and how active gel parameters control the architecture, orientation, connectivity and dynamics of self-organized patterns. Finally, we substantiate with discrete network simulations the main requirements for nematic bundle formation according to our theory, namely increased active tension perpendicular to the nematic direction and generalized active forces conjugate to nematic order. Our work portrays actin gels a reconfigurable active materials with a spontaneous tendency to develop patterns of dense nematic bundles.

Introduction

Actin networks are remarkably dynamic and versatile and organize in a variety of architectures to accomplish crucial cellular functions [1]. For instance, isotropic thin actin gels form the cell cortex, which largely determines cell shape [2] and motility in confined non-adherent environments [3, 4]. Polar structures at the edge of adherent cells, either forming filaments as in filopodia or sheets as in lamellipodia [5], enable cells to probe their environment and crawl on substrates. Nematic actin bundles conform a variety of contractile structures [6], including sub-cellular rings during cytokinesis [7] and cortical repair [8], supra-cellular rings during wound healing [9] or development [10], bundle networks during cellularization [11], or stress fibers in adherent cells [12–14]. Nematic bundles consist of highly aligned and densely packed actin filaments of mixed polarity connected by a diversity of crosslinkers. In vivo and in vitro observations show the key role of actin nucleators and regulators, of myosin activity and of crosslinkers in the assembly and maintenance of actin bundles [1, 2, 5, 6, 13, 15–21].

Various studies have emphasized the morphological, dynamical, molecular and functional specificities of different types of actin bundles such as dorsal, transverse, and ventral stress fibers or contractile rings [13, 14, 2224]. Here, we ask the question of whether, despite this diversity, the ubiquity of actin bundles in different contexts can be explained by the intrinsic ability of the active actomyosin gel to self-organize into patterns of dense nematic structures. Suggestive of such active self-organization, stress fibers often form dynamic highly organized patterns, e.g. involving families of fibers along orthogonal directions [12, 14, 19, 22, 25–27]. Other kinds of actin bundles also develop patterns of remarkable regularity. For instance, parallel arrangements of 2 pm-spaced actin bundles serve as templates for extra-cellular matrix deposition during butterfly wing morphogenesis and determine their iridescence [28]. Similarly, the morphogenesis of the striated tracheal cuticle in Drosophila is pre-patterned by a parallel arrangement of actin bundles spaced by ~1 μm, spanning from sub-cellular to supra-cellular and organ scales [21, 29]. Muscle-like actin bundles form regular parallel patterns spanning organs as in C elegans [19] or the entire organism of hydra [30]. Furthermore, dense nematic bundles have been shown to assemble de novo from the sparse isotropic cortex in a process controlled by myosin activity [20], and to form a mechanically integrated network with the cortex [31].

To understand the mechanisms underlying the selfassembly of dense actin bundles from a low-density isotropic cortex, we develop a theory for the selforganization of dense nematic structures in the actomyosin cytoskeleton based on a nematic active gel theory [7, 32, 33] but accounting for density variations and compressibility [34, 35]. In our theory, nematic patterning is driven by activity rather than by a more conventional crowding mechanism [36, 37]. Linear stability analysis and fully nonlinear simulations show that, when coupled to nematodynamics, the well-known patterning mechanism based on self-reinforcing flows [29, 38, 39] leads to a rich diversity of patterns combining density and nematic order. See [34] for a related study, further discussed later. The geometry and dynamics of the emergent patterns are very similar to those observed in diverse cellular contexts. Finally, we test the key requirements on phenomenological activity parameters for such self-organization according to our theory using discrete network simulations.

Theoretical model

We describe a thin layer actomyosin cytoskeleton with a 2D nematic active gel theory [35]. A number of active cellular and cytoskeletal systems have been modeled using nematic active liquid-crystal theory, including dense colonies of elongated cells or dense confined cytoskeletal gels [40–42]. In such systems, crowding drives strong nematic order everywhere except at defects, which are generated by activity or required by topology [37]. We argue that the porous actin cytoskeleton is not a nematic liquidcrystal because it can adopt extended isotropic/low-order phases, and hence defects are not topologically required. Furthermore, the main mechanism driving nematic order in actomyosin gels is active contractility [19, 20, 22]. Finally, liquid crystals are nearly incompressible, whereas actin gels develop large density variations. We thus consider an active gel model capable of accommodating large density variations and in which nematic ordering is actively driven.

At time t and position x, the local state of the system is described by the areal density of cytoskeletal material ρ(t, x) and by the network architecture given by the symmetric and traceless nematic tensor q(t, x), see Fig. 1a. Assuming a fixed volumetric density ρvol, p can be mapped to cytoskeletal thickness ρ/ρvol. The nematic tensor can be expressed as qij = S (ninj − δij/2), where n is the average molecular alignment, S=2qijqij the degree of local alignment about n and δij is the Kronecker delta; here and in the following we use Einstein summation convention for repeated indices. We denote by υ(t, x) the velocity field of the gel and by d=12(Δυ+ΔυT) and w=12(Δυ+ΔυT) the rate-of-deformation and spin tensors. The rate of change of q relative to a frame that translates and locally rotates with the flow generated by υ is given by the Jaumann derivative q^=∂q/∂t+υ⋅Δq−w⋅q+q⋅w [43].

Key model ingredients

(a) The local state is defined by areal density ρ and by orientational order quantified by the nematic parameter S and by the nematic direction n. (b) Isotropic active tension λ when the network is isotropic (S = 0), and (c) anisotropic tension when S ≠ 0, controlled by κ = λaniso/λ. Orientational order is driven by (d) active forces conjugate S and characterized by parameter λ⊙ and by (e) passive flow-induced alignment in the presence of deviatoric rate-of-deformation with coupling parameter β.

A systematic derivation of the governing equations summarized next is given elsewhere [35], based on On-sager’s variational formalism of irreversible thermodynamics. Because we consider a bi-periodic domain, we ignore other boundary conditions. Transport and turnover of the cortical material is described by ∂ρ∂t+∇.(ρυ)−DΔρ+kd(ρ−ρ0)=0, (1)

where D is an effective diffusivity, ρ0 is the steady-state areal density and kd is the depolymerization rate. Balance of linear momentum takes the form ργυ = ∇ · σ, where γ > 0 models friction with a background medium and σ is the Cauchy stress tensor, which in 2D has units of tension. This tensor is not symmetric and reads σij=ρ{ 2η(dij+dkkδij)+βq^ij+σijact−L[ ∇iqkl∇jqkl−1ρqik∇l(ρ∇lqjk)+1ρqjk∇l(ρ∇lqik) ] }, (2)

where η > 0 is the shear viscosity, β < 0 measures the dissipative coupling between nematic order and strain rate [7], which here induces a stress proportional to changes in nematic tensor, σijact is the active tension resulting from mechanical transduction of of chemical power in the gel, and L > 0 is the Frank constant. To model the dependence of contractility on network architecture, we assume σijact=λδij+λanisoqij=λ(δij+kqij), (3)

where κ = λaniso/λ measures the sign and strength of active tension anisotropy. When order is low (S ≈ 0), active tension is isotropic, Fig. 1b, whereas when order is high, active tension becomes anisotropic, Fig. 1c, with active tension along the nematic direction reflecting the sliding of antiparallel fibers driven by myosin motors, and perpendicular to it reflecting the the out-of-equilibrium binding of bundling proteins or myosins [5, 44–50].

Balance of the generalized forces power-conjugate to q^ also includes viscous, elastic-nematic and active contributions, and takes the form ηrotq^+βddev+(2a+bS2)q−L(Δq+∇q.∇ρρ)−ρλ⊙q=0, (4)

where ηrot is a nematic viscous coefficient, the second term models alignment induced by strain rate (Fig. 1e) with the deviatoric part of the rate-of-deformation tensor given by dijdev=dij−(dkk/2)δij , and a > 0 and b > 0 are susceptibility coefficients. The last term is an active generalized force controlled by activity parameter λ⊙ ≥ 0 tending to further align filaments (Fig. 1d) [7]. This term is linear in ρ because in the expansion λ¯⊙+ρλ⊙ the constant contribution λ¯⊙ can be subsumed by the susceptibility parameter a. Thus, the active term acts as a negative density-dependent susceptibility. When c0 = 2a–ρ0λ⊙ < 0, the system can sustain a uniform quiescent state with p(x, t) = ρ0, υ(x,t) = 0 and a non-zero nematic tensor satisfying S2 = −c0/b. Even if c0 > 0 and hence the uniform quiescent state is devoid of order, pattern formation can induce density variations such that 2a – ρλ⊙ becomes locally negative and actively favors local nematic order. The entropy production inequality requires that 2η ηrot − β2 ≤ 0 [35].

By freezing an isotropic state, S = 0, our model reduces to an orientation-independent active gel model, which develops periodic patterns driven by self-reinforcing active flows sustained by diffusion and turnover [29]. In the present model, however, translational, orientational and density dynamics are intimately coupled through the terms involving β, λ⊙, κ and L.

We readily identify the hydrodynamic length ℓs = η/γ , above which friction dominates over viscosity, the Damkolher length ℓD=D/kd above which reactions dominate over diffusion, and the nematic length ℓq=L/|2a−λ⊙ρ0 . Non-dimensional analysis reveals a set of non-dimensional groups that control the system behavior, namely the non-dimensional turnover rate k¯d=ℓs2/ℓD2 , the Frank constant L¯=L/(ηD) , the susceptibility parameters ā = a/(γD) and b¯=b/(γD) , the drag coefficients η¯rot=ηrot/η and β¯=β/η , the nematic activity coefficient λ¯⊙=ρ0λ⊙/(γD) , and the active tension parameters λ¯=λ/(γD) and κ (Section II of SI). The full list of material parameters for each figure given in Supplementary Tables (I,II) and justified in Section IV of SI.

Results

Onset and nature of pattern formation

To examine the role of nematic order in the emergence of various actin architectures, we performed linear stability analysis of our model particularized to 1D, whose dynamical variables are velocity, density and nematic order, υ(x, t), ρ(x,t), and q(x,t), along x, where q > 0 (< 0) corresponds to a nematic orientation n parallel (perpendicular) to the x–axis (Section I of SI). We first focused on the case c0 = 2a – ρ0λ⊙ > 0 to examine the loss of stability of a uniform, isotropic, and quiescent steady state (υ(x,t) = υ0, υ(x,t) = 0, q(x, t) = 0) by increasing the master activity parameter λ and identifying the most unstable modes. This allowed us to determine a threshold activity for pattern formation and the wavelength of the emerging pattern. Since the exact evaluation of such quantities requires solving nonlinear equations, we derived explicit expansions in the limit of small L for the critical contractile activity λcrit≈λcrit,0[ 1−12ℓsℓD(1+2ℓsℓD)δ ]+ο(δ2), (5)

where λcrit,0=(γD+2kdη)2=γD(1+2ℓs/ℓD)2 and δ=γDκβ/(2ηc0) , and for the the corresponding wavenumber νcrit2≈νcrit,02[ 1+18(1+2ℓsℓD)2δ ]+ο(δ2), (6)

where vcrit,02=[ kdγ/(4ηD) ]1/2=1/(2ℓsℓD) .

The active origin of pattern formation in our model is evident from these expressions. When κ = 0 or β = 0, and hence δ = 0, we recover the predictions of an active gel model not accounting for network architecture [29], λcrit = λcrit,0 and υcrit = υcrit,0. However, active tension anisotropy (κ ≠ 0) and flow-induced alignment (β < 0) fundamentally change the nature of pattern formation. Nematic order introduces quantitative changes in critical tension and wavenumber, which can be very significant depending on the ratio of hydrodynamic and Damkolher lengths and on the strength and sign of nematic coupling. The nematic corrections increase as c0 → 0, close to the point where the uniform quiescent state develops spontaneous order. We thus studied separately the regime 0 < c0 ≪ 1 (Section III of SI), finding analogous expansions for the critical tension and wavenumber in terms of δ = γDκβ/(2ηL). Interestingly, Eq. (5) shows that the activity threshold is reduced, and hence pattern formation facilitated, when κ < 0, i.e. when active tension is larger perpendicular to the nematic direction. Besides modifying critical tension and wavenumber, the present model predicts that the dynamical modes with self-reinforcing flows generate patterns where high density co-localizes with high nematic order.

To test the validity of this analysis and further understand the system beyond the onset of pattern formation, we performed fully non-linear finite element simulations in a periodic 2D domain [35]. In these simulations, we increased the activity parameter λ beyond the instability starting from a quiescent uniform state. We found that the linear stability analysis very accurately predicts the activity thresholds and pattern wave-numbers within two percent across a wide range of parameters. In the nonlinear regime, the exponentially-growing instabilities eventually reach out-of-equilibrium quasi-steadystate patterns involving self-reinforcing flows towards regularly-spaced regions of high density surrounded by a low density matrix. In the absence of nematic coupling (β = κ = λ⊙ = 0), these high density domains are droplets arranged in a regular hexagonal lattice with S = 0 throughout the domain, Movie 1. In contrast, for a generic parameter set with finite β, κ and λ⊙, high density domains adopt elongated configurations or bands where order is high, surrounded by a low density and low order matrix, Fig. 2b and Movie 2. Thus, the nematic active gel develops out-of-equilibrium localized states through a chemomechanical mechanism [51], which unlike those in [34], exhibit localization of both density and nematic order, and hence resemble dense nematic structures embedded in a low-density isotropic actin cortex. Furthermore, the shape and internal architecture of dense and nematic phases are qualitatively modified by the nematic coupling.

Active patterns coupling nematic order and density driven by self-reinforcing flows

(a) Dimensionless order parameter characterizing relative orientation of nematic direction and high-density bands given by ω=ℓs2/(ρ02|A|)∫A∇ρ⋅q∇ρ dS as a function of active tension anisotropy parameter κ, showing transition from states with nematic direction parallel to high-density structures (ω < 0, fibrillar patterns) for κ < 0 to states with nematic direction perpendicular to high-density structures (ω > 0, sarcomeric patterns) for κ > 0. (b) Map of density, nematic order S, nematic direction (red segments) and flow field (green arrows) for quasi-steady fibrillar (I) and sarcomeric (III) patterns, and for a transition pattern of high density droplets with high nematic order (II) corresponding to nearly isotropic active tension. (c) These quasi-steady states are out-of-equilibrium and maintained by self-reinforcing flows, diffusion and turnover.

Our simulations show that self-reinforcing flows develop along the direction of largest active tension, and consequently the pattern architecture depends on the sign of κ, Fig. 2c. For κ < 0, the system self-organizes into high-density and high-order bands, where nematic direction is parallel to their axis, in what we call fibrillar pattern, Fig. 2b(I). Instead, for κ > 0 nematic order is perpendicular to the axis of the bands, in what we call sarcomeric pattern, Fig. 2b(III). To systematically study the effect of active tension anisotropy, we varied K between -0.8 and 0.8 while keeping all other nondimensional groups fixed and setting λ to be 1.3 times the critical activity. We defined the order parameter ω, Fig. 2a, allowing us to distinguish between sarcomeric (ω > 0) and fibrillar (ω < 0) organizations. We found a sharp transition between fibrillar and sarcomeric regimes around κ ≈ 0, during which elongated high-density and high-order domains fragment into nematic droplets or tactoids [52, 53], Fig. 2b(II).

Our results for κ < 0, leading to self-organized dense nematic fibrillar patterns from an isotropic low-density network, are in agreement with evidence suggesting that stress fibers can assemble from the actin cortex without the involvement of stress fiber precursors or actin polymerization at focal adhesions [20]. They also agree with observations showing that actin bundles form a mechanical continuum with the surrounding sparse and isotropic cortex [31]. Their morphology and patterning dynamics is strikingly reminiscent of actin microridges, formed at the apical surfaces of mucosal epithelial cells [55, 56]. Finally, we also note the similarity in terms of density and nematic architecture between our fibrillar patterns and those emerging in other active systems through different mechanisms of self-organization, including polar motile filaments [57, 58] or mean-field models of dry mixtures of microtubules and motors [59].

Requirements for fibrillar and sarcomeric patterns

At linear order, our theory shows that the distinctly nematic self-organization requires both flow-induced alignment (β) and active tension anisotropy (κ), whereas no condition is required on nematic activity (λ⊙). We performed further simulations to establish the requirements for fibrillar and sarcomeric active patterning in the nonlinear regime. We found that both sar-comeric and fibrillar patterns readily form for β = 0 and finite κ, yet a finite value of β enhances fibrillar formation, leading to longer and more stable dense bands, and hinders sarcomeric organization, Movie 3. This behavior is expected since the velocity gradients of the selfreinforcing flows tend to align filaments parallel to high-density bands due to βddev in Eq. (4).

The active nematic coefficient λ⊙ modifies the onset of pattern formation through C0 according to linear stability analysis. Apart from this linear effect, it should also contribute to condensation in high-density regions since it appears multiplied by ρ in Eq. (4). To examine this nonlinear effect, we perform simulations with λ⊙ = 0 but keeping c0 fixed. This leads to very different patterns without clear elongated structures and very mild nematic patterning, Supplementary Figure 1(a). Enhancing nematic patterning by considering the largest thermodynamically allowed value of |β| leads to elongated structures for κ < 0, but rather than high-order co-localizing with high density, the nematic field develops domains with 90° angles between high- and low-density regions as in [34], Supplementary Figure 1(b), in an architecture enhanced by higher friction γ, Supplementary Figure 1(c). Thus, the architectures found for λ⊙ =0 and κ < 0 are distinct from the fibrillar pattern described previously. Similarly, rather than sarcomeres, for λ⊙ = 0, κ > 0 and high |β | we found patterns of nematic asters (high- density droplets with radial nematic organization around them), Supplementary Figure 1(b,c).

Together, these results show that active tension anisotropy (κ ≠ 0) and nematic activity (λ⊙ ≠ 0) are necessary and sufficient for nematic self-organization into fibrillar or sarcomeric patterns, with flow-induced alignment (β < 0) favoring fibrillar organization.

Morphological and dynamical diversity of selforganized fibrillar patterns

Given the morphological and dynamical diversity of nematic bundles in actin gels across cell types, geometric confinement, mechanical environment, or biological and pharmacological treatments [11, 26, 27, 60–62], we varied model parameters to examine the architectures predicted by our active gel model, focusing on κ < 0. Significant changes in the effective parameters of our active gel model are reasonable since the active mechanical properties of actomyosin gels strongly depend on micro-architecture both in reconstituted systems and in cells [49, 63].

With our default parameter set, bundle junctions and free ends are unfavorable and reorganize during pattern formation to annihilate as much as possible, Movie 4(I). However, because the initial state of the system is isotropic but fibrillar patterns are not, this coarsening process leads to frustrated labyrinth patterns with domains and defects, which depending on parameters can remain frozen in quasi-steady states as in Fig. 3a(I). We then asked the question of whether an orientational bias, which physically may be caused by cytoskeletal flows, boundaries or directed polymerization [22], could direct the pattern and result in fewer defects. We first slightly modified the system by including a small anisotropic strain-rate, according to which friction is computed relative to an elongating background. This directional bias is sufficient to produce well-oriented defect-free patterns aligned with the direction of elongation, Fig. 3a(II) and Movie 4(II). Alternatively, we considered c0 to be slightly negative, leading to the uniform and nematic steady state ρ(x,t) = ρ0, υ(x, t) = 0 and q(x,t)=q0=±12−c0/b . The linear stability analysis around this state and further nonlinear simulations show that the essential phenomenology of Eqs. (5,6) and Fig. 2 is not altered by the slight pre-existing order (Section III of SI). Again, the weak pre-existing order provides sufficient bias to direct pattern orientation, Fig. 3a(III). Thus, our model indicates that an anisotropic bias can guide and anneal nematic fibrillar patterns, in agreement with remarkably oriented patterns of actin bundles in elongated cells [28], as a result of uniaxial cellular stretch [19, 56], or on anisotropically curved surfaces [21, 29].

Control of nematic bundle pattern orientation, connectivity and dynamics

(a) Effect of orientational bias. (I) A uniform isotropic gel self-organizes into a labyrinth pattern with defects. (II) A small background anisotropic strain-rate efficiently orients nematic bundles. (III) A slight initial network alignment (S0 = 0.05) orients bundles, which later loose stability, bend, and generate/anneal defects. See Movie 4. (b) Depending on active tension anisotropy, nematic bundles are contractile and straighten (I, κ = -0.2), leading to quasi-steady networks, or extensile and wrinkle (II, κ = -0.8), leading to bundle breaking and recombination, and persistently dynamic networks (III). See Movie 5. (c) Promoting mechanical interaction between bundles. (I) Dynamical pattern obtained by reducing friction, and thereby increasing a¯,b¯,λ¯⊙ and k¯d . Time sequence in bottom indicates a typical reconfiguration event during which weak bundles (dashed) become strong ones (solid) and vice-versa. (II) Nearly static pattern obtained increasing k¯d , (III) which becomes highly dynamic by further increasing k¯d . Time sequence in bottom indicates illustrate the collapse (black), expansion (purple) and splitting (green) of cells in the network. See Movie 6. We indicate by * dynamical patterns exhibiting spatiotemporal chaos.

At longer times, the nematic bundles in Fig. 3a(III) develop secondary active instabilities leading to coordinated bending, curvature amplification, defect nucleation and annihilation, Movie 4(III), in a behavior reminiscent of active extensile systems. This possibility is puzzling because our active gel is contractile. To systematically examine it, we performed simulations at higher active tension anisotropies κ = −0.8, which we compared with our reference κ = −0.2, Fig. 3b. While in our reference system bundles behave like contractile objects tending to straighten, for κ = −0.8 they behave like extensile objects enhancing curvature, which results in continuous defect nucleation as highly bent bundles destabilize and fragment, as well as defect annihilation as pairs of free ends merge to reorganize the network, in a behavior akin to active turbulence [64]. See Movie 5 for an illustration and for the slightly extensile case κ = −0.5. To understand the origin of this behavior, we examined the individual components of the total tension, Eq. (2), along and perpendicular to the fibrillar pattern, Supplementary Fig. 2. Because in all of these simulations active contraction is larger perpendicular to nematic bundles (κ < 0) as required for their self-assembly, bundles are effectively extensile with regards to the active component of the total tension. Competing with active tension, however, viscous tension is negative and larger perpendicular to the bundles. Hence, depending on their relative magnitude, the total tension can be larger along or perpendicular to the bundles. We found that for contractile bundles leading to stable fibrillar patterns (κ = −0.2), total tension is larger along bundles, indicating that a large fraction of the active power perpendicular to the bundles is dissipated in the self-reinforcing flows, and hence is not available to perform power at a mesoscale. Instead, for stronger active anisotropy (κ = −0.8), total tension along nematic bundles is smaller than perpendicular to them, Supplementary Fig. 2(c), confirming the extensile nature of the self-organized bundles leading to the secondary wrinkling instability. In summary, our theory predicts that a contractile nematic active gel can self-organize into fibrillar patterns with mesoscale bundles that are either contractile or extensile depending on the parameter regime.

Focusing on contractile bundles, we then examined a different parameter regime known to trigger sustained pattern dynamics. For isotropic gels, previous work has shown that reducing friction triggers chaotic dynamics as the distance between high-density regions, 2π/υcrit, becomes comparable or smaller than the hydrodynamic length scale [29], thus enabling their mechanical interaction. In a model devoid of orientational order, reducing friction is equivalent to increasing k¯d . In our model, however, we can either reduce γ, which in non-dimensional terms means increasing k¯d , ā, b¯ and λ¯⊙ in concert, or increase k¯d while leaving all other non-dimensional parameters fixed. The first of these choices leads to dynamic and hierarchical networks with very dense and highly-aligned bundles, which coexist with perpendicular weak bundles with much lower density enrichment and ordering. These two families of bundles enclose cells of isotropic and sparse gel, Fig. 3c(I). Junctions where two or more dense bundles meet are very unfavorable and short-lived, but junctions of two dense and a weak bundle are much more stable. Because bundles are mechanically coupled, the networks actively reorganizes in events where dense bundles become weak bundles and vice-versa, inset and Movie 6(I). We note that here total tension along bundles is much larger than perpendicular to them, Supplementary Fig. 2(d), and hence the persistent dynamics are unrelated to the previously described behavior of extensile bundles.

The second choice to favor mechanical interaction of bundles, increasing k¯d only, leads to very different networks with high-density aster-like clusters interconnected by straight actin bundles. Because now λ¯⊙ is not particularly large, order is low at the core of these clusters, enabling high-valence networks where four bundles often meet at one cluster. For k¯d=10 , the network is stable and nearly crystalline, Fig. 3b(II), whereas for k¯d=20 , it becomes highly dynamical and pulsatile with frequent collapse of polygonal cells by fusion of neighboring actin clusters and their attached bundles (black polygons) and nucleation of new bundles within large low-density cells (dashed/solid green lines), Fig. 3b(III) and Movie 6(III). This architecture and dynamics resemble those of early C elegans embryos [65], adherent epithelial cells treated with epidermal growth factor [27] and mouse embryonic stem cells [61]. Recent active gel models accounting for RhoA signaling develop similar pulsatile behaviors in 2D, but do not predict the orientational order of the spatiotemporal patterns of the actomyosin cortex [66].

In summary, our theory maps how effective parameters of the actin gel control the active self-organization of a uniform and isotropic gel into a pattern of high-density nematic bundles embedded in a low-density isotropic matrix, including the activity threshold, the bundle spacing, orientation, connectivity and dynamics.

Microscopic origin of κ < 0 and λ⊙ > 0 through discrete network simulations

A somewhat counterintuitive prediction of our model is that the self-organization of nematic bundles, the most prominent emerging organization in actin gels across cell types and length-scales, requires that active tension perpendicular to nematic orientation is larger than along this direction (κ < 0), at least at the onset of pattern formation. While our continuum hydrodynamical model can address mesoscopic conditions for self-organization, it cannot provide insight about the microscopic origin of effective activity parameters. To examine whether the conditions κ < 0 and λ⊙ > 0 for spontaneous formation of fibrillar patterns are plausible from a microscopic point of view, we performed discrete network simulations using Cytosim [67].

Addressing the self-organization of the actin cytoskeleton at mesoscales directly with discrete network simulations is very challenging due to the wide range in time-and length-scales and the need to realistically model diffusion, network renewal, friction and gel hydrodynamics. Instead, we aimed at characterizing the out-ofequilibrium mechanical behavior of a representative volume element of the material with uniform mesoscopic properties, Fig. 4a. We performed 2D simulations in which semi-flexible filaments interact with crosslinkers and myosin motors, all of which undergo turnover and have a stoichiometry previously used to model the actin cytoskeleton [68], Fig. 4b. See Section V of SI for a detailed description of the simulation protocol. Briefly, we modified Cytosim to account for average orientational order in the simulation box, which we evaluated as a sample average of orientations over the ensemble of segments composing the filaments. We further introduced a nematic energy penalty in the network allowing us to restrain average nematic order to a target value S0.

Assessment of activity parameters κ and λ⊙ through discrete network simulations

(a) Illustration of the computational domain of the discrete network as a uniform representative volume element of the gel. (b) Sketch of model ingredients and setup to compute tension along and perpendicular to the nematic direction. (c) Typical time-signal for parallel and perpendicular tensions following addition of crosslinkers and motors (translucent lines) along with time average (solid lines) for isotropic and anisotropic networks. Tension is normalized by mean tension σ¯=(σ||+σ⊥)/2 computed from timeaverages and time by actin turnover time. (d) Mean tension as a function of network density for several nematic parameters S0, where both quantities are normalized by their values for the lowest density. With this normalization, Eq. (7) predicts a linear dependence with slope λ = 1 (dashed line). Error bars span two standard deviations. (e) Deviatoric tension as a function of nematic order for different densities. The dashed line is a linear regression to simulation data. (f) Dynamics of nematic order in a periodic network following addition of crosslinkers and motors for three initial values of nematic order. (g) Rate of change of nematic order normalized by turnover rate as a function of initial nematic order at zero and finite temperature.

We first prepared a system consisting only of randomly oriented actin fibers, imposed the desired orientational order S0 using the nematic penalty and equilibrated the system. In a first set of simulations, once S0 was reached, we deactivated the nematic penalty and added crosslinkers and myosins, driving the system out-of-equilibrium. The free contraction of the system was prevented by the addition of anchors at the boundary of the representative volume element, which also allowed us to compute anchor forces and hence estimate the effective active tension along the nematic direction σ|| = σijninj and perpendicular to it, σ⊥ = σijmimj with nimi =0 and mimi = 1, Fig. 4(b).

Addition of crosslinkers and myosins leads to bundling of actin filaments at the microscale, Movie 7, distinct from the mesoscale fibrillar pattern formation emerging from the active gel model. It also leads to out-ofequilibrium tension as measured by the anchors. For isotropic networks (S0 = 0), active tension is isotropic with σ|| ≈ σ⊥. For anisotropic networks, however, we found that tension becomes anisotropic with σ⊥ > σ||, Fig. 4(c).

We systematically characterized this behavior varying initial orientational order and network density. According to our active gel model, Eqs. (2,3), in the absence of nematic gradients and flow, the tension components σ|| and σ⊥ satisfy the following relations in terms of mean and deviatoric tensions σ¯=(σ∥+σ⊥)/2=λρ and (σ∥−σ⊥)/σ¯=κS. (7)

Remarkably, our discrete network simulations closely followed these relations, Fig. 4(d,e), which allowed us to estimate κ ≈ −1.6. We robustly found that κ < 0 for perturbations of selected parameters of the discrete network model as long as turnover rates of crosslinkers and myosins were relatively fast.

We then wondered if the discrete network simulations could provide evidence for the orientational activity parameter in our theory, ρλ⊙. For a uniform system with nematic order along a given direction and ignoring the susceptibility parameter b, Eq. (4) becomes ηrotS˙+(2a−ρλ⊙)S+Ks(S−S0)=0, (8)

where ηrot is the viscous drag of the filaments in the discrete network simulations, a > 0 the entropic tendency of the model to return to isotropy, ρλ⊙ the active forcing of nematic order resulting from crosslinkers and motors, and the last term accounts for the effect of the nematic penalty with coefficient KS. As a first test of this model, we started from an isotropic and periodic network and tracked the athermal dynamics of S under the action of the nematic penalty in the absence of anchors, crosslinkers and myosins. For λ⊙ = 0 and a = 0, Eq. (8) predicts an exponential relaxation given by S(t) = S0(1 − e−KSt/ηrot), which very closely matched the simulation data for different values of S0 and for a single fitting parameter ηrot, Supplementary Fig. 3(III) and Movie 8. We then deactivated the nematic penalty and added crosslinkers and motors, but not anchors, to track unconstrained dynamics of nematic order starting from different values of S0. In agreement with the notion of an active force driving nematic order, we found that S(t) monotonically increased, Fig. 4(f). More quantitatively, we tested the short-time prediction of Eq. (8), ηrotS = (ρλ⊙ − 2a)S0, by plotting S as estimated from our simulations, as a function of S0, Fig. 4(g). We found a nearly linear relation with positive slope, hence providing evidence for an active generalized force driving order. In agreement with the theory, in the athermal limit, the tendency to actively increase order is faster as the en-tropic tendency to disorder is absent (a = 0).

In summary, discrete network cytoskeletal simulations provide a microscopic justification for two key ingredients of our active gel theory, namely that nematic order elicits (1) anisotropic active tensions, which can be larger perpendicular to the nematic direction (κ < 0), and (2) active generalized forces driving further ordering.

Conclusions

We have developed a theory for the active selforganization of initially uniform and isotropic actin gels into various localized dense nematic architectures embedded in an isotropic matrix of low density. This model predicts a variety of emergent patterns involving asters, tactoids, and sarcomeric bands. More importantly, it identifies a wide parameter space where the active gel spontaneously develops patterns of dense nematic bundles, the most prominent nematic architecture across scales and cell types. We have characterized how the activity threshold, spacing, geometry, connectivity and dynamics these patterns depends on effective active gel parameters. Because these mesoscale parameters depend on the composition and dynamics of the network at the molecular scale, our results portray actin gels as responsive and reconfigurable active materials that cells can finely regulate. We have further shown that the spontaneous tendency of the gel to assemble bundle patterns can be directed via subtle cues. Thus, a combination of biochemical control of actin dynamics along with geometric, mechanical or biochemical guiding [69, 70] may explain the emergence and context-dependent organization of regular patterns of bundle networks, from sub-cellular to organism scales [12, 14, 19, 21, 22, 25–28, 30]. Consistently, perturbations of myosin contractility have been shown to alter, disorder, or even prevent the formation patterns of parallel bundles in C elegans [19], whereas perturbations of actin polymerization in drosophila embryos impair the robust organization of actin bundle patterns at the cellular and organ scales by disrupting orientation and spacing, but not the tendency of the actomyosin cytoskeleton to form patterns of parallel bundles [21].

Our theory identifies two key requirements on activity parameters for the self-organization of patterns of nematic bundles, namely active tension anisotropy with larger tension perpendicular to the nematic direction and generalized active forces tending to increase nematic order. Although our discrete network simulations support these two conditions, we expect that in a different regime anisotropic tension may be larger along the nematic direction (κ > 0). For instance, once bundles are dense and maximally aligned, the ability of the active nematic gel to perform active tension perpendicular to the nematic direction may saturate, while myosin motors may contract the gel along the nematic direction more effectively. The regime studied here explains the initial assembly of dense nematic bundles, but not their maturation to become highly contractile or viscoelastic as demonstrated for stress fibers depending on different isoforms of nonmuscle myosin II or on actin regulators such as zyxin [71, 72]. Our work thus suggests further experimental and computational work to establish a comprehensive mapping between molecular and mesoscale properties of the active gel, and hence a bridge between molecular regulation and emergent cytoskeletal organization.

eLife Assessment

In this study, the authors offer a theoretical explanation for the emergence of nematic bundles in the actin cortex, carrying implications for the assembly of actomyosin stress fibers. As such, the study is a valuable contribution to the field actomyosin organization in the actin cortex. While the theoretical work is solid, experimental evidence in support of the model assumptions remains incomplete. The presentation could be improved to enhance accessibility for readers without a strong background in hydrodynamic and nematic theories.

Reviewer #1 (Public Review):

Summary: In this article, Mirza et al developed a continuum active gel model of actomyosin cytoskeleton that account for nematic order and density variations in actomyosin. Using this model, they identify the requirements for the formation of dense nematic structures. In particular, they show that self-organization into nematic bundles requires both flow-induced alignment and active tension anisotropy in the system. By varying model parameters that control active tension and nematic alignment, the authors show that their model reproduces a rich variety of actomyosin structures, including tactoids, fibres, asters as well as crystalline networks. Additionally, discrete simulations are employed to calculate the activity parameters in the continuum model, providing a microscopic perspective on the conditions driving the formation of fibrillar patterns.

Strengths: The strength of the work lies in its delineation of the parameter ranges that generate distinct types of nematic organization within actomyosin networks. The authors pinpoint the physical mechanisms behind the formation of fibrillar patterns, which may offer valuable insights into stress fiber assembly. Another strength of the work is connecting activity parameters in the continuum theory with microscopic simulations.

Weaknesses: This paper is a very difficult read for nonspecialists, especially if you are not well-versed in continuum hydrodynamic theories. Efforts should be made to connect various elements of theory with biological mechanisms, which is mostly lacking in this paper. The comparison with experiments is predominantly qualitative. It is unclear if the theory is suited for in vitro or in vivo actomyosin systems. The justification for various model assumptions, especially concerning their applicability to actomyosin networks, requires a more thorough examination. The classification of different structures demands further justification. For example, the rationale behind categorizing structures as sarcomeric remains unclear when nematic order is perpendicular to the axis of the bands. Sarcomeres traditionally exhibit a specific ordering of actin filaments with alternating polarity patterns. Similarly, the criteria for distinguishing between contractile and extensile structures need clarification, as one would expect extensile structures to be under tension contrary to the authors' claim. Additionally, its unclear if the model's predictions for fiber dynamics align with observations in cells, as stress fibers exhibit a high degree of dynamism and tend to coalesce with neighboring fibers during their assembly phase. Finally, it seems that the microscopic model is unable to recapitulate the density patterns predicted by the continuum theory, raising questions about the suitability of the simulation model.

Reviewer #2 (Public Review):

Summary:

The article by Waleed et al discusses the self organization of actin cytoskeleton using the theory of active nematics. Linear stability analysis of the governing equations and computer simulations show that the system is unstable to density fluctuations and self organized structures can emerge. While the context is interesting, I am not sure whether the physics is new. Hence I have reservations about recommending this article.

Strengths:

(i) Analytical calculations complemented with simulations (ii) Theory for cytoskeletal network

Weaknesses:

Not placed in the context or literature on active nematics.

Reviewer #3 (Public Review):

The manuscript "Theory of active self-organization of dense nematic structures in the actin cytoskeleton" analysis self-organized pattern formation within a two-dimensional nematic liquid crystal theory and uses microscopic simulations to test the plausibility of some of the conclusions drawn from that analysis. After performing an analytic linear stability analysis that indicates the possibility of patterning instabilities, the authors perform fully non-linear numerical simulations and identify the emergence of stripe-like patterning when anisotropic active stresses are present. Following a range of qualitative numerical observations on how parameter changes affect these patterns, the authors identify, besides isotropic and nematic stress, also active self-alignment as an important ingredient to form the observed patterns. Finally, microscopic simulations are used to test the plausibility of some of the conclusions drawn from continuum simulations.

The paper is well written, figures are mostly clear and the theoretical analysis presented in both, main text and supplement, is rigorous. Mechano-chemical coupling has emerged in recent years as a crucial element of cell cortex and tissue organization and it is plausible to think that both, isotropic and anisotropic active stresses, are present within such effectively compressible structures. Even though not yet stated this way by the authors, I would argue that combining these two is of the key ingredients that distinguishes this theoretical paper from similar ones. The diversity of patterning processes experimentally observed is nicely elaborated on in the introduction of the paper, though other closely related previous work could also have been included in these references (see below for examples).

To introduce the continuum model, the authors exclusively cite their own, unpublished pre-print, even though the final equations take the same form as previously derived and used by other groups working in the field of active hydrodynamics (a certainly incomplete list: Marenduzzo et al (PRL, 2007), Salbreux et al (PRL, 2009, cited elsewhere in the paper), Jülicher et al (Rep Prog Phys, 2018), Giomi (PRX, 2015),...). To make better contact with the broad active liquid crystal community and to delineate the present work more compellingly from existing results, it would be helpful to include a more comprehensive discussion of the background of the existing theoretical understanding on active nematics. In fact, I found it often agrees nicely with the observations made in the present work, an opportunity to consolidate the results that is sometimes currently missed out on. For example, it is known that self-organised active isotropic fluids form in 2D hexagonal and pulsatory patterns (Kumar et al, PRL, 2014), as well as contractile patches (Mietke et al, PRL 2019), just as shown and discussed in Fig. 2. It is also known that extensile nematics, \kappa<0 here, draw in material laterally of the nematic axis and expel it along the nematic axis (the other way around for \kappa>0, see e.g. Doostmohammadi et al, Nat Comm, 2018 "Active Nematics" for a review that makes this point), consistent with all relative nematic director/flow orientations shown in Figs. 2 and 3 of the present work.

The results of numerical simulations are well-presented. Large parts of the discussion of numerical observations - specifically around Fig. 3 - are qualitative and it is not clear why the analysis is restricted to \kappa<0. Some of the observations resonate with recent discussions in the field, for example the observation of effectively extensile dynamics in a contractile system is interesting and reminiscent of ambiguities about extensile/contractile properties discussed in recent preprints (https://arxiv.org/abs/2309.04224). It is convincingly concluded that, besides nematic stress on top of isotropic one, active self-alignment is a key ingredient to produce the observed patterns.

I compliment the authors for trying to gain further mechanistic insights into this conclusion with microscopic filament simulations that are diligently performed. It is rightfully stated that these simulations only provide plausibility tests and, within this scope, I would say the authors are successful. At the same time, it leaves open questions that could have been discussed more carefully. For example, I wonder what can be said about the regime \kappa>0 (which is dropped ad-hoc from Fig. 3 onward) microscopically, in which the continuum theory does also predict the formation of stripe patterns - besides the short comment at the very end? How does the spatial inhomogeneous organization the continuum theory predicts fit in the presented, microscopic picture and vice versa?

Overall, the paper represents a valuable contribution to the field of active matter and, if strengthened further, might provide a fruitful basis to develop new hypothesis about the dynamic self-organisation of dense filamentous bundles in biological systems.

Author response:

The following is the authors’ response to the original reviews.

eLife assessment

In this study, the authors offer a theoretical explanation for the emergence of nematic bundles in the actin cortex, carrying implications for the assembly of actomyosin stress fibers. As such, the study is a valuable contribution to the field actomyosin organization in the actin cortex. While the theoretical work is solid, experimental evidence in support of the model assumptions remains incomplete. The presentation could be improved to enhance accessibility for readers without a strong background in hydrodynamic and nematic theories.

To address the weaknesses identified in this assessment, we have expanded the motivation and description of the theoretical model, specifically insisting on the experimental evidence supporting its rationale and assumptions. These changes in the revised manuscript are implemented in the two first paragraphs of Section “Theoretical model” and in a more detailed description and justification of the different mathematical terms that appear in that section. We have made an effort to map in our narrative different terms to mechanistic processes in the actomyosin network. Even if the nature of the manuscript is inevitably theoretical, we think that the revised manuscript will be more accessible to a broader spectrum of readers.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

In this article, Mirza et al developed a continuum active gel model of actomyosin cytoskeleton that account for nematic order and density variations in actomyosin. Using this model, they identify the requirements for the formation of dense nematic structures. In particular, they show that self-organization into nematic bundles requires both flow-induced alignment and active tension anisotropy in the system. By varying model parameters that control active tension and nematic alignment, the authors show that their model reproduces a rich variety of actomyosin structures, including tactoids, fibres, asters as well as crystalline networks. Additionally, discrete simulations are employed to calculate the activity parameters in the continuum model, providing a microscopic perspective on the conditions driving the formation of fibrillar patterns.

Strengths:

The strength of the work lies in its delineation of the parameter ranges that generate distinct types of nematic organization within actomyosin networks. The authors pinpoint the physical mechanisms behind the formation of fibrillar patterns, which may offer valuable insights into stress fiber assembly. Another strength of the work is connecting activity parameters in the continuum theory with microscopic simulations.

We thank the referee for these comments.

Weaknesses:

(A) This paper is a very difficult read for nonspecialists, especially if you are not well-versed in continuum hydrodynamic theories. Efforts should be made to connect various elements of theory with biological mechanisms, which is mostly lacking in this paper. The comparison with experiments is predominantly qualitative.

We understand the point of the referee. While it is unavoidable to present the continuum hydrodynamic theory behind our results, we have made an effort in the revised manuscript to (1) motivate the essential features required from a theoretical model of the actomyosin cytoskeleton capable of describing its nematic self organization (two first paragraphs of Section “Theoretical model”), and to (2) explicitly explain the physical meaning of each of the mathematical terms in the theory, and when appropriate, relate them to molecular mechanisms in the cytoskeleton. We hope that the revised manuscript addresses the concern of the referee.

Regarding the comparison with experiments, they are indeed qualitative because the main point of the paper is to establish a physical basis for the self-organization of dense nematic structures in actomyosin gels. Somewhat surprisingly, we argue that a compelling mechanism explaining the tendency of actomyosin gels to form patterns of dense nematic bundles has been lacking. As we review in the introduction, these patterns are qualitatively diverse across cell types and organisms in terms of geometry and dynamics, and for this reason, our goal is to show that the same material in different parameter regimes can exhibit such qualitative diversity. A quantitative comparison is difficult for several reasons. First, many of the parameters in our theory have not been measured and are expected to vary wildly between cell types. In fact, estimates in the literature often rely on comparison with hydrodynamic models such as ours. For this reason, we chose to delineate regimes leading to qualitatively different emerging architectures and dynamics. Second, the patterns of nematic bundles found across cell types depend on the interaction between (1) the intrinsic tendency of actomyosin gels to form such structures studied here and (2) other elements of the cellular context. For instance, polymerization and retrograde flow from the lamellipodium, the physical barrier of the nucleus, and the interaction with the focal adhesion machinery are essential to understand the emergence of stress fibers in adherent cells. Cell shape and curvature anisotropy control the orientation of actin bundles in parallel patterns in the wings and trachea of insects. Nuclear positions guide the actin bundles organizing the cellularization of Sphaeroforma arctica [11]. Here, we focus on establishing that actomyosin gels have an intrinsic ability to self organize into dense nematic bundles, and leave how this property enables the morphogenesis of specific structures for future work. We have emphasized this point in the revised section of conclusions.

(B) It is unclear if the theory is suited for in vitro or in vivo actomyosin systems. The justification for various model assumptions, especially concerning their applicability to actomyosin networks, requires a more thorough examination.

We thank the referee for this comment. Our theory is applicable to actomyosin gels originating from living cells. To our knowledge, the ability of reconstituted actomyosin gels from purified proteins to sustain the kind of contractile dynamical steady-states observed in living cells is very limited. In the revised manuscript, we cite a very recent preprint presenting very exciting but partial results in this direction [49]. Instead, reconstituted in vitro systems encapsulating actomyosin cell extracts robustly recapitulate contractile steady-states. This point has been clarified in the first paragraph of Section “Theoretical model”.

(C) The classification of different structures demands further justification. For example, the rationale behind categorizing structures as sarcomeric remains unclear when nematic order is perpendicular to the axis of the bands. Sarcomeres traditionally exhibit a specific ordering of actin filaments with alternating polarity patterns.

We agree with the referee and in the revised manuscript we have avoided the term “sarcomeric” because it refers to very specific organizations in cells. What we previously called “sarcomeric patterns”, where bands of high density exhibit nematic order perpendicular to the axis of the bands, is not a structure observed to our knowledge in cells. It is introduced to delimit the relevant region in parameter space. In the revised manuscript, we refer to this pattern as “banded pattern with perpendicular nematic organization” or “banded pattern” in short.

(D) Similarly, the criteria for distinguishing between contractile and extensile structures need clarification, as one would expect extensile structures to be under tension contrary to the authors' claim.

We thank the referee for raising this point, which was not sufficiently clarified in the original manuscript. We first note that in incompressible active nematic models, active tension is deviatoric (traceless and anisotropic) because an isotropic component would simply get absorbed by the pressure field enforcing incompressibility. Being compressible, our model admits an active tension tensor with deviatoric and isotropic components. We consider always a contractile (positive) isotropic component of active tension, but the deviatoric component can be either contractile (𝜅 > 0) or extensile (𝜅 < 0), where we follow the common terminology according to which in contractile/extensile active nematics the active stress is proportional to q with a positive/negative proportionality constant [see e.g. https://doi.org/10.1038/s41467018-05666-8]. Furthermore, as clarified in the revised manuscript, total active stresses accounting for the deviatoric and isotropic components are always contractile (positive) in all directions, as enforced by the condition |𝜅| < 1.

For fibrillar patterns, we need 𝜅 < 0, and therefore active stresses are larger perpendicular to the nematic direction. This means that the anisotropic component of the active tension is extensile, although, accounting for the isotropic component, total active tension is contractile (see Fig. 1c). This is now clarified in the text following Eq. 7 and in Fig. 1.

However, following fibrillar pattern formation and as a result of the interplay between active and viscous stresses, the total stress can be larger along the emergent dense nematic structures (“contractile structures”) or perpendicular to them (“extensile structures”). To clarify this point, in the revised Fig. 4 and the text referring to it, we have expanded our explanation and plotted the difference between the total stress component parallel to the nematic direction (𝜎∥) and the component perpendicular to the nematic direction (𝜎⊥), with contractile structures satisfying 𝜎∥ − 𝜎⊥ > 0 and extensile structures satisfying 𝜎∥ − 𝜎⊥ < 0. See lines 280 to 303. This is consistent with the common notion of contractile/extensile systems in incompressible nematic systems [see e.g. https://doi.org/10.1038/s41467-018-05666-8].

(E) Additionally, its unclear if the model's predictions for fiber dynamics align with observations in cells, as stress fibers exhibit a high degree of dynamism and tend to coalesce with neighboring fibers during their assembly phase.

In the present work, we focus on the self-organization of a periodic patch of actomyosin gel. However, in adherent cells boundary conditions play an essential role, as discussed in our response to comment (A) by this referee. In ongoing work, we are studying with the present model the dynamics of assembly and reconfiguration of dense nematic structures in domains with boundary conditions mimicking in adherent cells, possibly interacting with the adhesion machinery, finding dynamical interactions as those suggested by the referee. As an example, we show a video of a simulation where at the edge of the circular domain, there is an actin influx modeling the lamellipodium, and in four small regions friction is higher simulating focal adhesions. Under these boundary conditions, the model presented in the paper exhibits the kind of dynamical reorganizations alluded by the referee.

Author response video 1.

We would like to note, however, that the prominent stress fibers in cells adhered to stiff substrates, so abundantly reported in the literature, are not the only instance of dense nematic actin bundles. In the present manuscript, we emphasize the relation of the predicted organizations with those found in different in vivo contexts not related to stress fibers, such as the aligned patterns of bundles in insects (trachea, scales in butterfly wings), in hydra, or in reproductive organs of C elegans; the highly dynamical network of bundles observed in C elegans early embryos; or the labyrinth patters of micro-ridges in the apical surface of epidermal cells in fish.

(F) Finally, it seems that the microscopic model is unable to recapitulate the density patterns predicted by the continuum theory, raising questions about the suitability of the simulation model.

We thank the referee for raising this question, which needs further clarification. The goal of the microscopic model is not to reproduce the self-organized patterns predicted by the active gel theory. The microscopic model lacks essential ingredients, notably a realistic description of hydrodynamics and turnover. Our goal with the agent-based simulations is to extract the relation between nematic order and active stresses for a small homogeneous sample of the network. This small domain is meant to represent the homogeneous active gel prior to pattern formation, and it allows us to substantiate key assumptions of the continuum model leading to pattern formation, notably the dependence of isotropic and deviatoric components of the active stress on density and nematic order (Eq. 7) and the active generalized stress promoting ordering.

We should mention that reproducing the range of out-of-equilibrium mesoscale architectures predicted by our active gel model with agent-based simulations seems at present not possible, or at least significantly beyond the state-of-the-art. To our knowledge, these models have not been able to reproduce the heterogeneous nonequilibrium contractile states involving sustained self-reinforcing flows underlying the pattern formation mechanism studied in our work. The scope of the discrete network simulations has been clarified in lines 340 to 349 in the revised manuscript.

While agent-based cytoskeletal simulations are very attractive because they directly connect with molecular mechanisms, active gel continuum models are better suited to describe out-of-equilibrium emergent hydrodynamics at a mesoscale. We believe that these two complementary modeling frameworks are rather disconnected in the literature, and for this reason, we have attempted substantiate some aspects of our continuum modeling with discrete simulations. We have emphasized the complementarity of the two approaches in the conclusions.

Reviewer #1 (Recommendations For The Authors):

Questions on the theory:

Does rho describe the density of actin or myosin? The authors say that they are modeling actomyosin material as a whole, but the actin and myosin should be modeled separately. Along, similar lines, does Q define the ordering of actin or myosin?

Active gel models of the actomyosin cytoskeleton have been formulated with independent densities for actin and for myosin or using a single density field, implicitly assuming a fixed stoichiometry. Super-resolution imaging of the actomyosin cytoskeleton also suggest that in principle it makes sense to consider different nematic fields for actin and for myosin filaments. In the revised manuscript, we now explicitly mention that our density and nematic field are effective descriptions of the entire actomyosin gel (lines 82-84).

A more detailed model would entail additional material parameters, not available experimentally, which may help reproduce specific experiments but that would make the systematic study of the different behaviors much more difficult. Our approach has been to keep the model minimal meeting the fundamental requirements outlined in the first paragraphs of Section “Theoretical model”.

Should the active stress depend on material density? It seems strange (from Eq. 3) that active stress could be non-zero even where density is zero, since sigma_act does not depend on rho.

Yes, active stress is assumed to be proportional to density. Eq. 3 in the original manuscript was misleading (it was multiplied by rho in Eq. 2). In the revised manuscript, we have explained with a bit more detail the theoretical model, clarifying this point.

The authors should clearly explain their rationale for retaining certain types of nonlinear terms while ignoring others in theory. For instance, the nonlinearities in the equations of motion are sometimes quadratic in the fields, while there are also some cubic terms. Please remark up to what order in the fields the various interactions are modeled.

We thank the referee for raising this point. The nonlinearities in the theory are easily explained on the basis of a small number of choices. We have added a new paragraph towards the end of Section “Theoretical model” (lines 145 to 152) providing a rationale for the origin and underlying assumptions leading to different nonlinearities.

To connect with experiments and the biological context, please explain the biological origin of various terms in the model: (1) L-dependent terms in Eq. 2 and 4, (2) Flowalignment of nematic order and experimental evidence in support of it, (3) densitydependent susceptibility terms in Eq. 4

(1) Unfortunately, the L-dependent terms are very bulky, but are very standard in nematic theories. The best way to understand their physical significance is through the expression of the nematic free-energy, which is now given and explained in the revised manuscript (Eq. 3). The resulting complicated expression for the molecular field and the nematic stress (Eqs. 4 and 5) are mathematical consequences of the choice of nematic free energy. In the revised manuscript, we also attempt to provide a basis for these terms in the context of the actin cytoskeleton. (2) To our knowledge, the best reference supporting this term from experiments is Reymann et al, eLife (2016). In the revised manuscript, we have provided a physical interpretation. (3) We have expanded the motivation and plausible microscopic justification of this term.

There are different 'activity' terms in the model. Their biophysical origin is not made clear. For example, the authors should make clear if these activities arise from filament or motor activity. Relatedly, the authors should provide a comprehensive discussion of the signs of the different active parameters and their physical interpretations.

In an active gel model, activity parameters are phenomenological and how they map to molecular mechanisms is not precisely known, although conventionally contractile active tension is ascribed to the mechanical transduction of chemical power by myosin motors. The fact is that, besides myosin activity, there are many nonequilibrium processes in the actomyosin cytoskeleton that may lead to active stresses including (de)polymerization of filaments or (un)binding of crosslinkers. In the revised manuscript, we have added sentences illustrating how different terms may result from microscopic mechanisms, but providing a precise mapping between our model and nonequilibrium dynamics of proteins is beyond the scope of our work, although our discrete network simulations address this issue to a certain degree.

Following the suggestion of the referee, our description of the theory now discusses much more extensively the signs of activity parameters and their physical interpretations, e.g. the text following Eq. 7.

Throughout the paper, various activity terms are varied independently of each other. Is that a reasonable assumption given that activities should depend on ATP and are thus not independent of one another?

We agree that, ultimately, all active process depend on the conversion of chemical energy into mechanical energy. However, recent work has highlighted how active tension also depends on the microscopic architecture of the network controlled by multiple regulators of the actomyosin cytoskeleton (e.g. Chug et al, Nat Cell Biol, 2017). It is reasonable to expect that, for a given rate of ATP consumption, chemical power will be converted into mechanical power in different ways depending on the micro-architecture of the cytoskeleton, e.g. the stoichiometry of filaments, crosslinkers, myosins, or the length distribution of filaments (very long filaments crosslinked by myosins may be difficult to reorient but may contract efficiently).

We have added a paragraph in Section “Theoretical model” with a discussion, lines 153 to 156.

Sarcomeres are muscle fibers that exhibit alternating polarity pattern. Such patterning is not evident in what the authors call 'sarcomeres' in Fig. 2. I believe the authors should revise their terminology and not loosely interpret existing classifications in the field.

We thank the referee for raising this point. We have changed the terminology.

Fig 2a: Is the cartoon for filament alignment incorrect for kappa>0?

The cartoon is correct. In the revised manuscript we have explained more clearly the physical meaning of kappa in the text following Eq. 7. In the caption of Fig. 1 and of Fig. 2a, we have also clarified that when the absolute value of kappa is <1, then active tension is positive in all directions.

Within the section "Requirements for fibrillar and banded patterns", it will be useful to show the figures for varying the different active parameters in the main figures.

We have followed the referee’s suggestion and moved Supp. Fig. 1 of the original manuscript to the main figures.

How do the authors decide if bundles are contractile or extensile? Why are contractile bundles under tension while extensile bundles are under compression? I would expect the opposite.

We agree that this point deserves a more detailed explanation. In the revised manuscript and in the new Figure 4, we further develop this point. The fibrillar pattern forms when kappa<0. We further assume that -1<kappa<0, so that active tension is positive in all directions. In this regime, the deviatoric (anisotropic) part of active tension is extensile. However, following pattern formation and because of the interplay between active and viscous stresses, the total stress in the emerging bundles may become extensile or contractile, depending on whether the largest component of stress is perpendicular or along the bundle axis. This is now presented in the updated figure, with new panels presenting maps of the total tension. The text discussing this point has been rewritten and we hope that the new version is much clearer (lines 280 to 303).

A contractile bundle tends to shorten, but it cannot do it because of boundary conditions or the interaction with other bundles. As a result they are in tension. Conversely, an extensile bundle tries to elongate, but being constrained, it becomes compressed. As an analogy, consider the cortex of a suspended cell. The cortex is contractile, but it cannot contract because of volume regulation in th cell, which is typically pressurized. As a result, tension in the cortex is positive, as shown by Laplace’s law [10.1016/j.tcb.2020.03.005]. We have tried to clarify this point in the revised manuscript.

Can the authors reproduce alternating density patterns using the cytosim simulations? This is an important step in establishing the correspondence between the continuum theory and the agent-based model.

We have addressed this point in our response to public comment (F) of this referee.

The authors do not provide code or data.

The finite element code with an input file require to run a representative simulation in the paper is now made available, see Ref. [74].

The customizations of Cytosim needed to account for nematic order in our discrete network simulations are available, see Ref. [98].

Reviewer #2 (Public Review):

Summary:

The article by Waleed et al discusses the self organization of actin cytoskeleton using the theory of active nematics. Linear stability analysis of the governing equations and computer simulations show that the system is unstable to density fluctuations and self organized structures can emerge. While the context is interesting, I am not sure whether the physics is new. Hence I have reservations about recommending this article.

We thank the referee for these comments. In the revised manuscript, we have highlighted the novelty, particularly in the last paragraph of the introduction, the first two paragraphs of Section “Theoretical model”, and in the conclusions. Despite a very large literature on theoretical models of stress fibers, actin rings, and active nematics, we argue that the active self-organization of dense nematic structures from an isotropic and low-density gel has not been compellingly explained so far. Many models assume from the outset the presence of actin bundles, or explain their formation using localized activity gradients. The literature of active nematics has extensively studied symmetry breaking and the self-organization. However, most of the works assume initial orientational order. Only a few works study the emergence of nematic order from a uniform isotropic state, but consider dry systems lacking hydrodynamic interactions or incompressible and density-independent systems [37,38]. Yet, pattern formation in actomyosin gels is characterized by large density variations, and by highly compressible flows, which coordinate in a mechanism relying on an advective instability and self-reinforcing flows.

Our theoretical model is not particularly novel, and as we mention in the manuscript, it can be particularized to different models used in the literature. However, we argue that it has the right minimal features to capture nematic self-organization in actomyosin gels. To our knowledge, no previous study explains the emergence of dense and nematic structures from a low-density isotropic gel as a result of activity and involving the advective instability typical of symmetry-breaking and patterning in the actomyosin cytoskeleton. These are important qualitative features of our results that resonate with a large experimental record, and as such, we believe that our work provides a new and compelling mechanism relying on self-organization to explain the prominence and diversity of patterns involving dense nematic bundles in the actomyosin cytoskeleton across species.

Strengths:

(i) Analytical calculations complemented with simulations (ii) Theory for cytoskeletal network

Weaknesses:

Not placed in the context or literature on active nematics.

We agree with the referee that this was a weakness of the original manuscript. In the revised manuscript, within reasonable space constraints given the size and dynamism of the field of active nematics, we have placed our work in the context of this field (end of introduction and first two paragraphs of Section “Theoretical model”). The published version of our companion manuscript [45] also contributes to providing a clear context to our theoretical model within the field.

Reviewer #2 (Recommendations For The Authors):

The article by Waleed et al discusses the self organization of actin cytoskeleton using the theory of active nematics. Linear stability analysis of the governing equations and computer simulations show that the system is unstable to density fluctuations and self organized structures can emerge. While the context is interesting, I am not sure whether the physics is new. Hence I have reservations about recommending this article. I explain my questions comments below.

We have responded to this comment above.

(i) Active nematics including density variations have been dealt quite extensively in the literature. For example, the works of Sriram Ramaswami have dealt with this system including linear stability analysis, simulations etc. In what way is the present work different from the system that they have considered?

(ii) Active flows leading to self organization has been a topic of discussion in many works. For example: (i) Annual Review of Fluid Mechanics, Vol. 43:637-659, 2010, https://doi.org/10.1146/annurev-fluid-121108-145434 (ii) S Santhosh, MR Nejad, A Doostmohammadi, JM Yeomans, SP Thampi, Journal of Statistical Physics 180, 699-709 (iii) M. G. Giordano1, F. Bonelli2, L. N. Carenza1,3, G. Gonnella1 and G. Negro1, Europhysics Letters, Volume 133, Number 5. In what way this work is different from any of these?

(iii) I am confused about the models used in the paper. There is significant literature from Prof. Mike Cates group, Prof. Julia Yeomans group, Prof. Marchetti's group who all use similar governing equations. In the present paper, I find it hard to understand whether the model used is similar to the existing ones in literature or are there significant differences. It should be clarified.

Response to (i), (ii) and (iii).

We completely agree with this referee (and also the previous referee), that the contextualization of our work in the field of active nematics was very insufficient. In the revised manuscript, the last paragraph of the introduction and the first two paragraphs of Section “Theoretical model” now address this point. In short, previous active nematic models predicting patterns with density variations have been either for dry active matter (disregarding hydrodynamic interactions), or for suspensions of active particles moving in an incompressible flow. None of these previous works predict nematic pattern formation as a result of activity relying on the advective instability and self-reinforcing compressible flows, leading to high density and high order bundles surrounded by an isotropic low density phase. Yet, these are fundamental features observed in actomyosin gels. Many works deal with symmetry-breaking of a system with pre-existing order, but very few address how order emerges actively from an isotropic state. We thank the referee for pointing at the paper by Santhosh et al, who nicely make this argument and is now cited. Our mechanism is fundamentally different from that in Santhosh, whose model is incompressible and ignores density variations.

We hope that the revised manuscript addresses this important concern.

(i) >(iv) Below Eqn 6, it starts by saying that the “...origin..is clear...” Its not. I don't understand the physical origin of the instability, and this should be clarified, may be with some illustrations.

We apologize for this unfortunate sentence, which we have rewritten in the revised manuscript (lines 181 to 185).

Reviewer #3 (Public Review):

The manuscript "Theory of active self-organization of dense nematic structures in the actin cytoskeleton" analysis self-organized pattern formation within a two-dimensional nematic liquid crystal theory and uses microscopic simulations to test the plausibility of some of the conclusions drawn from that analysis. After performing an analytic linear stability analysis that indicates the possibility of patterning instabilities, the authors perform fully non-linear numerical simulations and identify the emergence of stripelike patterning when anisotropic active stresses are present. Following a range of qualitative numerical observations on how parameter changes affect these patterns, the authors identify, besides isotropic and nematic stress, also active self-alignment as an important ingredient to form the observed patterns. Finally, microscopic simulations are used to test the plausibility of some of the conclusions drawn from continuum simulations.

The paper is well written, figures are mostly clear and the theoretical analysis presented in both, main text and supplement, is rigorous. Mechano-chemical coupling has emerged in recent years as a crucial element of cell cortex and tissue organization and it is plausible to think that both, isotropic and anisotropic active stresses, are present within such effectively compressible structures. Even though not yet stated this way by the authors, I would argue that combining these two is of the key ingredients that distinguishes this theoretical paper from similar ones. The diversity of patterning processes experimentally observed is nicely elaborated on in the introduction of the paper, though other closely related previous work could also have been included in these references (see below for examples).

We thank the referee for these comments and for the suggestion to emphasize the interplay of isotropic and anisotropic active tension, which is possible only in a compressible gel, as mentioned in the revised manuscript. We have emphasized this point in different places in the revised manuscript. We thank the suggestions of the referee to better connect with existing literature.

To introduce the continuum model, the authors exclusively cite their own, unpublished pre-print, even though the final equations take the same form as previously derived and used by other groups working in the field of active hydrodynamics (a certainly incomplete list: Marenduzzo et al (PRL, 2007), Salbreux et al (PRL, 2009, cited elsewhere in the paper), Jülicher et al (Rep Prog Phys, 2018), Giomi (PRX, 2015),...). To make better contact with the broad active liquid crystal community and to delineate the present work more compellingly from existing results, it would be helpful to include a more comprehensive discussion of the background of the existing theoretical understanding on active nematics. In fact, I found it often agrees nicely with the observations made in the present work, an opportunity to consolidate the results that is sometimes currently missed out on. For example, it is known that self-organised active isotropic fluids form in 2D hexagonal and pulsatory patterns (Kumar et al, PRL, 2014), as well as contractile patches (Mietke et al, PRL 2019), just as shown and discussed in Fig. 2. It is also known that extensile nematics, \kappa<0 here, draw in material laterally of the nematic axis and expel it along the nematic axis (the other way around for \kappa>0, see e.g. Doostmohammadi et al, Nat Comm, 2018 "Active Nematics" for a review that makes this point), consistent with all relative nematic director/flow orientations shown in Figs. 2 and 3 of the present work.

We thank the referee for these suggestions. Indeed, in the original submission we had outsourced much of the justification of the model and the relevant literature to a related pre-print, but this is not reasonable. The companion publication has now been accepted in the New Journal of Physics, with significant changes to better connect the work to the field of active nematics. A preprint reflecting those changes is available in Ref. [64], but we hope to reference the published paper that will come out soon.

In the revised manuscript, we have significantly rewritten the Section “Theoretical model” to frame the continuum model in the context of the field of active nematics. While our model and results have commonalities with previous work, there are also important differences. We have highlighted the novelty of the present work along with the relation with previous studies and theoretical models in the last paragraph of the introduction and the first two paragraphs of Section “Theoretical model”. Furthermore, as suggested by the referee, we have made an effort to connect our results with previous work by Kumar, Mietke, Doostmohammadi and others.

Regarding the last point alluded by the referee (“extensile nematics, \kappa<0 here, draw in material laterally of the nematic axis and expel it along the nematic axis”), the picture raised by the referee would be nuanced for our compressible system as compared to the incompressible systems discussed in that reference. As we have elaborated in our response to point (D) of Referee #1, our systems are overall contractile (with positive active tension in all directions), but the deviatoric component of the active tension can be either extensile or contractile. In our “extensile” models (left in Fig. 2c), material is drawn to laterally to the nematic axis but it is not expelled along this axis. Instead, it is “expelled” by turnover. In the revised manuscript, we have added a comment about this.

The results of numerical simulations are well-presented. Large parts of the discussion of numerical observations - specifically around Fig. 3 - are qualitative and it is not clear why the analysis is restricted to \kappa<0. Some of the observations resonate with recent discussions in the field, for example the observation of effectively extensile dynamics in a contractile system is interesting and reminiscent of ambiguities about extensile/contractile properties discussed in recent preprints (https://arxiv.org/abs/2309.04224). It is convincingly concluded that, besides nematic stress on top of isotropic one, active self-alignment is a key ingredient to produce the observed patterns.

We thank the referee for these comments. We are reluctant to extend the detailed analysis of emergent architectures and dynamics to the case \kappa > 0 as it leads to architectures not observed, to our knowledge, in actin networks. In the revised manuscript, we have expanded and clarified the characterization of emergent contractile/extensile networks by reporting the relative magnitude of stress along and perpendicular to the nematic direction. Our revised manuscript clearly shows that even though all of our simulations describe locally contractile systems with extensile anisotropic active tension, the emergent meso-structures can be either extensile or contractile, with the extensile ones exhibiting the usual bend-type instability (a secondary instability in our system) described classically for extensile active nematic systems. We have rewritten the text discussing this (lines 280 to 303), where we have placed these results in the context of recent work reporting the nontrivial relation between the contractility/extensibility of the local units vs the nematic pattern.

I compliment the authors for trying to gain further mechanistic insights into this conclusion with microscopic filament simulations that are diligently performed. It is rightfully stated that these simulations only provide plausibility tests and, within this scope, I would say the authors are successful. At the same time, it leaves open questions that could have been discussed more carefully. For example, I wonder what can be said about the regime \kappa>0 (which is dropped ad-hoc from Fig. 3 onward) microscopically, in which the continuum theory does also predict the formation of stripe patterns - besides the short comment at the very end? How does the spatial inhomogeneous organization the continuum theory predicts fit in the presented, microscopic picture and vice versa?

We thank the referee for this compliment. We think that the point raised by the referee is very interesting. It is reasonable to expect that the sign of \kappa may not be a constant but rather depend on S and \rho. Indeed, for a sparse network with low order, the progressive bundling by crosslinkers acting on nearby filaments is likely to produce a large active stress perpendicular to the nematic direction, whereas in a dense and highly ordered region, myosin motors are more likely to effectively contract along the nematic direction whereas there is little room for additional lateral contraction by additional bundling. As discussed in our response to referee #1, we believe that studying the formation of patterns using the discrete network simulations is far beyond the scope of our work. We discuss in lines 332 to 341, as well as in the last paragraph of the conclusions, the scope and limitations of our discrete network simulations.

Overall, the paper represents a valuable contribution to the field of active matter and, if strengthened further, might provide a fruitful basis to develop new hypothesis about the dynamic self-organisation of dense filamentous bundles in biological systems.

Reviewer #3 (Recommendations For The Authors):

The statement "the porous actin cytoskeleton is not a nematic liquid-crystal because it can adopt extended isotropic/low-order phases" is difficult to understand and should be clarified, as the next paragraph starts formulating a nematic active liquid crystal theory. Do the authors mean a crystal that "Tends to be in a disordered phase?", according to its equilibrium properties? It would still be a "nematic liquid crystal", only its ground state is not a nematic phase.

We agree with the referee, and we hope that changes in the introduction and in Section “Theoretical model” address this comment.

I could not find what Frank energy is precisely used, that would be helpful information.

In the revised manuscript, we have provided the expression for the nematic free energy in Eq. 3.

The Significance of green/purple arrows in Fig 2a sketch unclear, green arrows also in b,c, do they represent the same quantity? From the simulations images it is overall it is very difficult to see how the flows are oriented near the high-density regions (i.e. if they are towards / away from the strip).

We thank the referee for bringing this up. The colorcodings of the sketches were confusing. The modified figures (Fig. 1(c) and Fig. 2(a)) present now a clearer and unified representation of anisotropic tension. The green arrows in Fig. 2(c) represent the out-of-equilibrium flows in the steady state. We agree that the zoom is insufficient to resolve the flow structure. For this reason, in the revised Fig. 2, we have added additional panels showing the flow with higher resolution.

It is currently unclear how the linear stability results - beyond identification of the parameter \delta - inform any of the remaining manuscript. Quantitative comparisons of the various length scales seen in simulated patterns (e.g. Fig. 2b, 3c etc) with linear predictions and known characteristic length scales would be instructive mechanistically, would make the overall presentation more compelling and probes limitations of linear results.

In the revised manuscript, we have provided further information so that the readers can appreciate the predictions and limitations of the linear stability results. We have added a sentence and a Figure to show that, in addition to the critical activity, the linear theory provides a good prediction of the wavelengh of the pattern. See lines 199 to 201.

It is not clear what is meant by "[bundle-formation] requires that active tension perpendicular to nematic orientation is larger than along this direction", and therefore also not why that would be "counter-intuitive". If interpreted naively, I would say that a large tension brings in more filaments into the bundle, so that may well be an obviously helpful feature for bundle formation and maintenance. In any case, it would be helpful if clarity is improved throughout when arguments about "directions of tensions" are made.

We have significantly rewritten the first paragraphs of section “Microscopic origin…” to clarify this point (lines 330 to 339). This paragraph, along with other changes in the manuscript such as the explanation of Eq. 7 or the discussion about the stress anisotropy in the new version of Fig. 4 (see lines 280 to 303), provide a better explanation of this important point.

All density color bars: Shouldn't they rather be labelled \rho/\rho_0?

Yes! We have corrected this typo.

Scalar product missing in caption definition of order parameter Fig. 2

We have corrected this typo.

Fig. 3a: I suggest to put the expression for q0 in the caption

We have changed q_0 by S_0 and clarified its meaning in the caption of what now is Fig 4.

Paragraph on bottom right of page 6 should several times probably refer to Fig. 3c(...), instead of Fig. 3b

We have corrected this typo.

Version 2

Abstract

The actin cytoskeleton is remarkably adaptable and multifunctional. It often organizes into nematic bundles such as contractile rings or stress fibers. However, how a uniform and isotropic actin gel self-organizes into dense nematic bundles is not fully understood. Here, using an active gel model accounting for nematic order and density variations, we identify an active patterning mechanism leading to localized dense nematic structures. Linear stability analysis and nonlinear finite element simulations establish the conditions for nematic bundle self-assembly and how active gel parameters control the architecture, orientation, connectivity and dynamics of self-organized patterns. Finally, we substantiate with discrete network simulations the main requirements for nematic bundle formation according to our theory, namely increased active tension perpendicular to the nematic direction and generalized active forces conjugate to nematic order. Our work portrays actin gels a reconfigurable active materials with a spontaneous tendency to develop patterns of dense nematic bundles.

Introduction

Actin networks are remarkably dynamic and versatile and organize in a variety of architectures to accomplish crucial cellular functions [1]. For instance, isotropic thin actin gels form the cell cortex, which largely determines cell shape [2] and motility in confined non-adherent environments [3, 4]. Polar structures at the edge of adherent cells, either forming filaments as in filopodia or sheets as in lamellipodia [5], enable cells to probe their environment and crawl on substrates. Nematic actin bundles conform a variety of contractile structures [6], including sub-cellular rings during cytokinesis [7] and cortical repair [8], supra-cellular rings during wound healing [9] or development [10], bundle networks during cellularization [11], or stress fibers in adherent cells [12–14]. Nematic bundles consist of highly aligned and densely packed actin filaments of mixed polarity connected by a diversity of crosslinkers. In vivo and in vitro observations show the key role of actin nucleators and regulators, of myosin activity and of crosslinkers in the assembly and maintenance of actin bundles [1, 2, 5, 6, 13, 15–21].

Various studies have emphasized the morphological, dynamical, molecular and functional specificities of different types of actin bundles such as dorsal, transverse, and ventral stress fibers or contractile rings [13, 14, 22–24]. Here, we ask the question of whether, despite this diversity, the ubiquity of actin bundles in different contexts can be explained by the intrinsic ability of the active actomyosin gel to self-organize into patterns of dense nematic structures. Suggestive of such active self-organization, stress fibers often form dynamic highly organized patterns, e.g. involving families of fibers along orthogonal directions [12, 14, 19, 22, 25–27]. Other kinds of actin bundles also develop patterns of remarkable regularity. For instance, parallel arrangements of 2 μm-spaced actin bundles serve as templates for extracellular matrix deposition during butterfly wing morphogenesis and determine their iridescence [28]. Similarly, the morphogenesis of the striated tracheal cuticle in Drosophila is pre-patterned by a parallel arrangement of actin bundles spaced by ~1μm, spanning from sub-cellular to supra-cellular and organ scales [21, 29]. Muscle-like actin bundles form regular parallel patterns spanning organs as in C. elegans [19] or the entire organism of hydra [30]. Furthermore, dense nematic bundles have been shown to assemble de novo from the sparse isotropic cortex in a process controlled by myosin activity [20], and to form a mechanically integrated network with the cortex [31].

Theoretical models of stress fibers often assume from the outset the existence of dense actin bundles, possibly embedded in an isotropic network [32], although previous work has studied the emergence of nematic rings as a result of localized activity gradients [33]. In the field of active nematics, many theoretical studies have examined the well-known hydrodynamical bend/splay instability arising in uniform active systems with pre-existing long-range orientational order [34–36]. However, the question of how patterns of orientational order arise from an isotropic and uniform system as a result of activity has received very little attention, and has been based so far on models that do not capture fundamental physical features of actomyosin gels as discussed below [37, 38]. Here, to understand the mechanisms underlying the self-assembly of dense actin bundles from a low-density isotropic cortex, we develop a theory for the self-organization of dense nematic structures in the actomyosin cytoskeleton based on a nematic active gel theory accounting for density variations and compressibility. In our theory, nematic patterning is driven by activity rather than by a more conventional crowding mechanism. Linear stability analysis and fully nonlinear simulations show that, when coupled to nematodynamics, the well-known patterning mechanism based on self-reinforcing flows [29, 39–43] leads to a rich diversity of patterns combining density and nematic order. The geometry and dynamics of the emergent patterns are very similar to those observed in diverse cellular contexts. Finally, we test key assumptions of our phenomenological theory leading to such self-organization using discrete network simulations.

Theoretical Model

The actomyosin cytoskeleton can be understood as a compressible, active and viscous fluid gel with orientational order undergoing turnover [44, 45]. Symmetry-breaking and pattern formation in actomyosin gels is often mediated by the emergence of an advective instability leading to compressible self-reinforcing flows. According to this mechanism, fluctuations in the density of cytoskeletal active units generate gradients in active stress, driving flows, which in turn advect the active units reinforcing the initial fluctuation [29, 39–41]. This kind of advective instability has been invoked to explain cell polarization and amoeboid motility [4, 41, 46], the formation of the cytokinetic ring [42], the formation of periodic dense actin structures during tracheal morphogenesis in Drosophila [29], or the emergence of self-sustained dynamical states in actomyosin gels extracted from cells [47, 48], and has been recently reproduced to some degree in confined reconstituted gels from purified proteins [49]. In all of these examples, the actomyosin gel develops sustained compressible flows converging towards regions of high density. Furthermore, the compressive strain rate induced by these active flows has been shown to drive nematic order [7, 33]. Finally, observations on adherent cells show that active contractility is required for actin bundle formation [19, 20, 22]. Therefore, a theoretical model to understand the self-organization of dense nematic structures in actomyosin gels should consider a compressible and density-dependent fluid capturing the advective instability mentioned above. Furthermore, this model should acknowledge the active nature of the assembly of dense nematic structures, and permit extended isotropic phases commonly observed in the actin cortex, possibly coexisting with dense nematic phases [20, 31], rather than thermodynamically enforce high nematic order everywhere except at topological defects [36, 50–52].

Previous models for dry and dilute aligning active matter develop density patterns [37, 53, 54] but fail to capture the hydrodynamic interactions of actomyosin gels, whereas models for active nematic fluids either ignore density [38, 55–61], or account for the density of active particles suspended in an incompressible flow [34, 35, 62, 63], and therefore cannot describe the advective instability and self-reinforcing flows of actomyosin gels. Previous models describing the emergence of nematic patterns from uniform and isotropic states either ignore hydrodynamics [37], or the density and flow compressibility characteristic of actomyosin gels [38], and therefore result in very different instability mechanisms to those presented here. To model a thin layer of actomyosin gel, we summarize next a minimal theory for 2D density-dependent compressible active nematic fluids. In [64], we provide a systematic derivation of this theory based on a variational formalism of irreversible thermodynamics. This model can be understood as a density-dependent and compressible version of the active nematic theory presented in [57], or as a nematic generalization of the isotropic theory used in [29]. As elaborated in [64], it is possible to develop an alternative compressible and density-dependent active nematic theory based on the Beris-Edwards formalism [38, 51, 55, 56, 58, 59, 63].

In our model, the local state of the system is described by the areal density of cytoskeletal material ρ(t, x) and by the network architecture given by the symmetric and traceless nematic tensor q(t, x), see Fig. 1a, both of which depend on time t and position x. A more detailed model could consider separate densities for actin, myosin, and possibly other structural or regulatory proteins. Likewise, in principle, the orientational order of actin and myosin filaments could be described by separate nematic tensors. The nematic tensor can be expressed as qij = S (ninj − δij/2), where n is the average molecular alignment, S=2qijqij the degree of local alignment about n and δij is the identity. We denote by v(t, x) the velocity field of the gel. The rate-of-deformation tensor d=12(∇υ+∇υT) measures the local rate of distortion of the fluid, whereas w=12(∇υ−∇υT) measures its local spin, where ∇ is the nabla differential operator. The deviatoric part of the rate-of-deformation tensor is defined by dijdev=dij−(dkk/2)δij . The rate of change of q relative to a frame that translates and locally rotates with the flow generated by υ is given by the Jaumann derivative q^=∂q/∂t+υ⋅∇q−w⋅q+q⋅w[50] .

Because we consider a bi-periodic domain Ω, we ignore other boundary conditions. Balance of cytoskeletal mass for a compressible fluid undergoing turnover takes the conventional form [29] ∂ρ∂t+∇.(ρυ)−DΔρ+kd(ρ−ρ0)=0, (1)

where the second term models advection of cytoskeletal material by flow, the third term models diffusion with D being an effective diffusivity, and the last term models cytoskeletal turnover, where ρ0 is the steady-state areal density and kd is the depolymerization rate. Note that for a uniform, steady-state and quiescent gel, the first three terms vanish and ρ(t, x) = ρ0.

Key model ingredients

(a) The local state is defined by areal density ρ and by orientational order quantified by the nematic parameter S and by the nematic direction n. (b) Isotropic active tension λ when the network is isotropic (S = 0), and (c) anisotropic tension when S ≠ 0, controlled by κ. Active tension is positive (contractile) in all directions whenever |κ| < 1, but its deviatoric part is contractile when κ > 0 and extensile when κ < 0. Orientational order is driven by (d) active forces conjugate to nematic order and characterized by parameter λ⊙ and by (e) passive flow-induced alignment in the presence of deviatoric rate-of-deformation with coupling parameter β.

Force balance in the gel can be expressed as ργυ=∇⋅σ, (2)

where γ > 0 models a viscous drag with the surroundings (e.g. the plasma membrane) and σ = σanem + σdiss + σact is the stress tensor in the gel, which in 2D has units of tension and which includes a contribution coming from the nematic free-energy, a dissipative contribution, and an active contribution.

The nematic stress σnem follows from a standard derivation adapted here to a density-dependent material. It derives from the free-energy ö = ∫Ωρf(q, ∇q)dS, where f(q,∇q)=12aS2+18bS4+12L|∇q|2 (3)

is the classical Landau expansion of free-energy density per unit mass, with a and b > 0 susceptibility parameters, and L > 0 the Frank constant. When a < 0, the first term favors equilibrium nematic ordering, e.g. due to crowding in very dense gels of elongated filaments. Otherwise, the susceptibility terms entropically favor isotropic states with small S. In the actin cytoskeleton, this can model the random orientation of filaments as a result of the entropic fluctuations of filaments and their nucleators. The last term penalizes sharp gradients in the nematic field, which can result from the bending rigidity of actin filaments. A lengthy but standard calculation leads to the so-called molecular field [64] hij=−δℱδqij=−ρ(2a+bS2)qij+L∇k(ρ∇kqij), (4)

and to the explicit form of the nematic non-symmetric stress tensor σijnem=−ρ∂f∂∇jqlk∇iqlk+qikjhjk−qjkjhik=L[−ρ∇iqkl∇jqkl+qik∇l(ρ∇lqjk)−qjk∇l(ρ∇lqik)]. (5)

The dissipative part of the stress σijdiss=ρ[2η(dij+dkkδij)+βq^ij], (6)

includes a viscous stress controlled by the gel viscosity parameter η [33] and a stress resulting from changes in the nematic field controlled by the coupling parameter η < 0 [7]. The term involving η can be understood as a stress in the gel arising from the drag induced by filaments as they reorient relative to the underlying hydrodynamic flow. Finally, we assume that the active stress resulting from the mechanical transduction of chemical power has an isotropic component and an anisotropic component oriented along the nematic tensor following σijact=ρ(λδij+λanisoqij)=ρλ(δij+κqij). (7)

Activity parameter λ controls the isotropic tension and is contractile for λ > 0, as assumed here. The additional activity parameter λaniso, or equivalently κ = λaniso/λ, controls the deviatoric (traceless) component of active tension. This component is anisotropic and can be positive or negative parallel or perpendicular to the nematic direction depending on the sign of λ. When |λ| < 1, then the total active tension remains positive in all directions, with a larger magnitude parallel to the nematic direction when λ > 0 (contractile deviatoric component) and perpendicular to it when λ < 0 (extensile deviatoric component). We note that the isotropic component of active tension is meaningful here because our active gel is compressible. When order is low (S ≈ 0), active tension is isotropic, Fig. 1b, whereas when order is high, active tension becomes anisotropic, Fig. 1c. We can interpret that active tension along the nematic direction reflects the sliding of antiparallel fibers driven by myosin motors, and whereas active tension perpendicular to it reflects the the out-of-equilibrium binding of bundling proteins or myosins [5, 65–71].

Balance of the generalized forces power-conjugate to q^ also includes viscous, elastic-nematic and active contributions, and takes the form ηrotq^+βddev−1ρh−ρλ⊙q=0. (8)

In this expression, ηrot is a nematic viscous coefficient. The second term models alignment induced by the rate of deformation of the flow, e.g. with compression/extension driving alignment perpendicular/parallel to the velocity gradient (Fig. 1e), as experimentally observed in [7]. This term involves the same coefficient β as the last term in Eq. (6) because of Onsager’s reciprocity relations, and the entropy production inequality requires that 2η ηrot − β2 ≥ 0 [64]. The third term is a thermodynamic force driven by F. In agreement with the observations that nematic ordering in actomyosin gels is actively driven, we assume a > 0, and therefore Eqs. (4,8) show that this term tends to restore isotropy. The last term is an active generalized force controlled by activity parameter λ⊙ ≥ 0 tending to further align filaments (Fig. 1d) [7]. This term is linear in ρ because in the expansion λ¯⊙+ρλ⊙ the constant contribution λ¯⊙ can be subsumed by the susceptibility parameter a [33]. Thus, the active term acts as a negative density-dependent susceptibility. When c0 = 2a − ρ0λ⊙ < 0, the system can sustain a uniform quiescent state with ρ(x, t) = ρ0, υ(x, t) = 0 and a non-zero nematic order parameter S02=−c0/b . Even if c0 > 0, and hence the uniform quiescent state is devoid of order, pattern formation can induce density variations such that 2a − ρλ⊙ becomes negative locally and actively favors local nematic order. Physically, the term −ρλ⊙q implements the notion that the binding of a bundling protein, which drives active alignment, is more probable when two filaments are in close proximity and nearly aligned, a situation favored by high density and nematic order.

The nonlinearity of the coupled system of partial differential equations given by the balance laws Eqs. (1,7,8), along with the constitutive relations in Eqs. (4,5,6,7), has different sources summarized next. The theory presents nonlinearities intrinsic to transport equations in the advective term of Eq. (1) and in the definition of the Jaumann derivative of the nematic tensor. Furthermore, nonlinearities in q in the constitutive relations results from the standard nematic free-energy adopted here. Our hypothesis that material properties in the gel are proportional to density and our thermodynamically-consistent derivation of the theory [64] result in further nonlinearities involving density in the constitutive relations. Finally, the nonlinearity involving density and the nematic field in the last term of Eq. (8) has been discussed in the previous paragraph.

Our theory has three active parameters, λ, κ and λ⊙, all reflecting the conversion of chemical power into mechanical power in the network. The magnitude and the modes of chemomechanical transduction should depend on the molecular architecture of the network [72, 73], e.g. the stoichiometry of filaments, cross-linkers, and myosins, or the length distribution of filaments. Accordingly, we allow these parameters to vary independently.

By freezing an isotropic state, S = 0, our model reduces to an orientation-independent active gel model, which develops periodic patterns driven by self-reinforcing active flows sustained by diffusion and turnover [29]. In the present model, however, translational, orientational and density dynamics are intimately coupled through the terms involving β, λ⊙, κ and L.

We readily identify the hydrodynamic length ls=η/γ , above which friction dominates over viscosity, the Damkolher length lD=D/kd above which reactions dominate over diffusion, and the nematic length lq = L|2a−λ⊙ρ0 . Non-dimensional analysis reveals a set of non-dimensional groups that control the system behavior, namely the non-dimensional turnover rate k¯d=ls2/lD2 , the Frank constant L¯=L/(ηD) , the susceptibility parameters a¯=a/(γD) and b¯=b/(γD) , the drag coefficients η¯rot=ηrot/η and β¯=β/η , the nematic activity coefficient λ¯⊙=ρ0λ⊙/(γD) , and the active tension parameters λ¯=λ/(γD) and κ (Appendix B). The full list of material parameters for each figure given in Tables I and II and justified in Appendix D.

Results

Onset and nature of pattern formation

To examine the role of nematic order in the emergence of various actin architectures, we performed linear stability analysis of our model particularized to 1D, whose dynamical variables are velocity, density and nematic order, v(x,t), ρ(x, t), and q(x,t), along x, where q > 0 (< 0) corresponds to a nematic orientation n parallel (perpendicular) to the x-axis (Appendix A). We first focused on the case c0 = 2a − ρ0λ⊙ > 0 to examine the loss of stability of a uniform, isotropic, and quiescent steady state (ρ(x, t) = ρ0, v(x,t) = 0, q(x,t) = 0) by increasing the master activity parameter λ and identifying the most unstable modes. This allowed us to determine a threshold activity for pattern formation and the wavelength of the emerging pattern (Appendix C). Since the exact evaluation of such quantities requires solving nonlinear equations, we derived explicit expansions in the limit of small L for the critical contractile activity λcrit≈λcrit,0[1−12ℓsℓD(1+2ℓsℓD)δ]+𝒪(δ2), (9)

where λ¯crit,0=(γD+2kdη)2=γD(1+2ℓs/ℓD)2 and δ = γDκβ/(2ηc0), and for the the corresponding wavenumber νcrit2≈νcrit,02[1+18(1+2ℓsℓD)2δ]+𝒪(δ2), (10)

where vcrit,02=[kdγ/4ηD]1/2=1/(2ℓsℓD) .

When κ = 0 or β = 0, and hence δ = 0, we recover the predictions of an active gel model not accounting for network architecture [29], λcrit = λcrit,0 and λcrit = λcrit, 0. The expression for λcrit,0 shows that the instability takes place when activity is large enough to overcome the effect of diffusion and turnover, tending to uniformize density, and that of friction and viscosity, tending to suppress flows. Active tension anisotropy (κ ≠ 0) and flow-induced alignment (β < 0) couple the instability of [29] to nematic order, changing the nature of pattern formation, see Eqs. (9,10) and the expression for δ. This leads to quantitative changes in critical tension and wavenumber, which can be very significant depending on the ratio of hydrodynamic and Damkolher lengths and on the strength and sign of nematic coupling. The nematic corrections increase as c0 → 0, close to the point where the uniform quiescent state develops spontaneous order. We thus studied separately the regime 0 < c0 ≪ 1, finding analogous expansions for the critical tension and wavenumber in terms of δ = γDκβ/(2ηL). Interestingly, Eq. (9) shows that the activity threshold is reduced, and hence pattern formation facilitated, when κ < 0, i.e. when active tension is larger perpendicular to the nematic direction. Besides modifying critical tension and wavenumber, the present model predicts that the dynamical modes with self-reinforcing flows generate patterns where high density co-localizes with high nematic order.

To test the validity of this analysis and further understand the system beyond the onset of pattern formation, we performed fully non-linear finite element simulations in a periodic square 2D domain [64, 74], with a domain size chosen to be 8 ℓpattern where ℓpattern = 2π/vcrit is the pattern lengthscale predicted by linear stability analysis. In these simulations, we increased the activity parameter A beyond the instability starting from a quiescent uniform state. We found that the linear stability analysis very accurately predicts the activity thresholds within two percent across a wide range of parameters. Furthermore, we found that the linear estimate for the wavenumber in Eq. (10) characterizes well the nonlinear patterns, as quantified in Supp. Fig. 1 and illustrated visually by the density patterns in Figs. 2 and 4 showing the entire 8ℓpattern × 8ℓpattern computational domain.. In the nonlinear regime, the exponentially-growing instabilities eventually reach out-of-equilibrium quasi-steady-state patterns involving self-reinforcing flows towards regularly-spaced regions of high density surrounded by a low density matrix, as previously reported for various compressible and density-dependent active gel models [29, 39–43]. In the absence of nematic coupling (β = κ = λ⊙ = 0), these high density domains are droplets arranged in a regular hexagonal lattice with S = 0 throughout the domain as previously reported [40], Movie 1. In contrast, for a generic parameter set with finite β, κ and λ⊙, high density domains adopt elongated configurations or bands where order is high, surrounded by a low density and low order matrix, Fig. 2b and Movie 2. Thus, our nematic active gel develops out-of-equilibrium localized nematic states starting from an isotropic state through a mechanochemical mechanism, which unlike those in [37, 38], exhibit localization of both density and nematic order, and hence resemble dense nematic structures embedded in a low-density isotropic actin cortex. Furthermore, the shape and internal architecture of dense and nematic phases are qualitatively modified by the nematic coupling.

Our simulations show that self-reinforcing flows develop along the direction of largest active tension, and consequently the pattern architecture depends on the sign of κ, Fig. 2c. For κ < 0, the system self-organizes into high-density and high-order bands, where nematic direction is parallel to their axis, in what we call fibrillar pattern, Fig. 2b(I). Instead, for κ > 0 nematic order is perpendicular to the axis of the bands, in what we call banded pattern, Fig. 2b(III). To systematically study the effect of active tension anisotropy, we varied κ between −0.8 and 0.8 while keeping all other non-dimensional groups fixed and setting λ to be 1.3 times the critical activity. We defined the order parameter ω, Fig. 2a, allowing us to distinguish between banded (ω > 0) and fibrillar (ω < 0) organizations. We found a sharp transition between fibrillar and banded regimes around κ ≈ 0, during which elongated high-density and high-order domains fragment into nematic droplets or tactoids [75, 76], Fig. 2b(II).

Active patterns coupling nematic order and density driven by self-reinforcing flows.

(a) Illustration of the dimensionless parameters characterizing active tension anisotropy (κ) and pattern architecture, quantified by the relative orientation of nematic order and high-density structures (ω=ℓs2/(ρ02|A|)∫A∇ρ⋅q⋅∇ρ dS) . (b) Order parameter of pattern architecture ω as a function of active tension anisotropy κ obtained from nonlinear simulations, showing transition from states with nematic direction parallel to high-density structures (ω < 0, fibrillar patterns) for κ < 0 to states with nematic direction perpendicular to high-density structures (ω > 0, banded patterns with perpendicular nematic organization) for κ > 0. Because |κ| < 1, the active tension is always positive in all directions. (c) Map of density, nematic order S, nematic direction (red segments) and flow field (green arrows) for quasi-steady fibrillar (I) and banded (III) patterns, and for a transition pattern of high density droplets with high nematic order (II) corresponding to nearly isotropic active tension. (d) Illustration of the out-of-equilibrium quasi-steady states, maintained by self-reinforcing flows, diffusion and turnover.

Our results for κ < 0, leading to self-organized dense nematic fibrillar patterns from an isotropic low-density network, are in agreement with evidence suggesting that stress fibers can assemble from the actin cortex without the involvement of stress fiber precursors or actin polymerization at focal adhesions [20]. They also agree with observations showing that actin bundles form a mechanical continuum with the surrounding sparse and isotropic cortex [31]. Their morphology and patterning dynamics is strikingly reminiscent of actin microridges, formed at the apical surfaces of mucosal epithelial cells [77, 78]. As discussed earlier, the condition κ < 0 implies that the deviatoric component of active tension is extensile. In agreement with widely studied incompressible extensile nematic systems, here the material is drawn perpendicular to the nematic direction, but rather than being expelled along the nematic direction as required from incompressibility [36], here it is recycled by disassembly and turnover (Fig. 2c,d left and Eq. (1)). Finally, we note the similarity in terms of density and nematic architecture between our fibrillar patterns and those emerging in other active systems through different mechanisms of self-organization, including polar motile filaments [79, 80] or mean-field models of dry mixtures of microtubules and motors [81].

Requirements for fibrillar and banded patterns

At linear order, our theory shows that the distinctly nematic self-organization requires both flow-induced alignment (β) and active tension anisotropy (κ), whereas no condition is required on nematic activity (λ⊙). We performed further simulations to establish the requirements for fibrillar and banded active patterning in the nonlinear regime. We found that both banded and fibrillar patterns readily form for β = 0 and finite κ, yet a finite value of β enhances fibrillar formation, leading to longer and more stable dense bands, and hinders banded organization, Movie 3. This behavior is expected since the velocity gradients of the self-reinforcing flows tend to align filaments parallel to high-density bands due to βddev in Eq. (8).

Pattern formation for a range of values of anisotropic active parameter κ in the limit λ⊙ → 0.

(a) Pattern formation for material parameters used in Fig. 2 except for λ⊙ = 0 while leaving c0 unchanged. (b) Here, in addition to λ⊙ = 0, we set β2 = 2ηηrot to the largest value allowed by the entropy production inequality. (c) Same parameters as in (b), except for an increase in friction as detailed in Table I.

The active nematic coefficient λ⊙ modifies the onset of pattern formation through c0 according to linear stability analysis. Apart from this linear effect, it should also contribute to condensation of nematic order in high-density regions since it appears multiplied by p in Eq. (8). To examine this nonlinear effect, we perform simulations with λ⊙ = 0 but keeping c0 fixed. This leads to very different patterns without clear elongated structures and very mild nematic patterning, Fig. 3a. Enhancing nematic patterning by considering the largest thermodynamically allowed value of |β| leads to elongated structures for κ < 0, but rather than high-order co-localizing with high density, the nematic field develops domains with 90° angles between high- and low-density regions, Fig. 3b, as observed in other active nematic systems [37]. This architecture is enhanced by higher friction γ, Fig. 3c. Thus, the architectures found for λ⊙ = 0 and κ < 0 are distinct from the fibrillar pattern described previously. Similarly, rather than bands with order perpendicular to their axis, for λ⊙ = 0, κ > 0 and high |β| we found patterns of nematic asters (high-density droplets with radial nematic organization around them), Fig. 3b,c.

Together, these results show that active tension anisotropy (κ ≠ 0) and nematic activity (λ⊙ = 0) are necessary and sufficient for nematic self-organization into fibrillar or banded patterns, with flow-induced alignment (β < 0) favoring fibrillar organization. Furthermore, they show that while Eqs. (9,10) accurately predict the activity threshold and the wavenumber of the nonlinear patterns, the linear stability analysis does not capture the architecture of the patterns, nor the key role of λ⊙ , in the nonlinear regime.

Morphological and dynamical diversity of self-organized fibrillar patterns

Given the morphological and dynamical diversity of nematic bundles in actin gels across cell types, geometric confinement, mechanical environment, or biological and pharmacological treatments [11, 26, 27, 82–84], we varied model parameters to examine the architectures predicted by our active gel model, focusing on κ < 0. Significant changes in the effective parameters of our active gel model are reasonable since the active mechanical properties of actomyosin gels strongly depend on micro-architecture both in reconstituted systems and in cells [70, 72].

With our default parameter set, bundle junctions and free ends are unfavorable and reorganize during pattern formation to annihilate as much as possible, Movie 4(I). However, because the initial state of the system is isotropic but fibrillar patterns are not, this coarsening process leads to frustrated labyrinth patterns with domains and defects, which depending on parameters can remain frozen in quasi-steady states as in Fig. 4a(I). We then asked the question of whether an orientational bias, which physically may be caused by cytoskeletal flows, boundaries or directed polymerization [22], could direct the pattern and result in fewer defects. We first slightly modified the system by including a small anisotropic strain-rate, according to which friction is computed relative to an elongating background. This directional bias is sufficient to produce well-oriented defect-free patterns aligned with the direction of elongation, Fig. 4a(II) and Movie 4(II). Alternatively, we considered c0 to be slightly negative, leading to the uniform and nematic steady state ρ(x,t) = ρ0, v(x,t) = 0 and q(x,t)=q0=±12−c0/b . The linear stability analysis around this state and further nonlinear simulations show that the essential phenomenology of Eqs. (9,10) and Fig. 2 is not altered by the slight pre-existing order (Appendix C). Again, the weak pre-existing order provides sufficient bias to direct pattern orientation, Fig. 4a(III). Thus, our model indicates that an anisotropic bias can guide and anneal nematic fibrillar patterns, in agreement with remarkably oriented patterns of actin bundles in elongated cells [28], as a result of uniaxial cellular stretch [19, 78], or on anisotropically curved surfaces [21, 29].

Control of nematic bundle pattern orientation, connectivity and dynamics.

(a) Effect of orientational bias. (I) A uniform isotropic gel self-organizes into a labyrinth pattern with defects. (II) A small background anisotropic strain-rate efficiently orients nematic bundles. (III) A slight initial network alignment (S0 = 0.05) orients bundles, which later loose stability, bend, and generate/anneal defects. See Movie 4. We recall that the nematic order parameter in the quiescent and uniform equilibrium state is S0 = 0 if c0 ≥ 0 and S0=−c0/b otherwise, with c0 = 2a − ρ0λ⊙. (b) Depending on active tension anisotropy, nematic bundles are contractile and straighten (I, κ = −0.2), leading to quasi-steady networks, or extensile and wrinkle (II, κ = −0.8), leading to bundle breaking and recombination, and persistently dynamic networks (III). See Movie 5. The contractility or extensibility of the nematic bundles can be appreciated in the maps of the difference between the stress along the nematic direction (σ||) and the stress perpendicular to it (σ⊥), normalized by the constant σcrit = λcritρ0. (c) Promoting mechanical interaction between bundles. (I) Dynamical pattern obtained by reducing friction, and thereby increasing a¯,b¯,λ¯⊙ and k¯d . Time sequence in the bottom indicates a typical reconfiguration event during which weak bundles (dashed) become strong ones (solid) and vice-versa. (II) Nearly static pattern obtained increasing k¯d , (III) which becomes highly dynamic by further increasing k¯d . Time sequence in the bottom indicates illustrate the collapse (black), expansion (purple) and splitting (green) of cells in the network. See Movie 6. We indicate by * dynamical patterns exhibiting spatiotemporal chaos.

At longer times, the nematic bundles in Fig. 4a(III) develop secondary active instabilities leading to coordinated bending, curvature amplification, defect nucleation and annihilation, Movie 4(III), in a behavior reminiscent of active extensile systems [34–36]. This possibility is far from obvious because, even though the deviatoric part of active tension is extensile since κ < 0, total active tension is contractile since |κ| < 1, Fig. 1c. To systematically examine this point, we performed simulations at higher active tension anisotropies κ = −0.8, which we compared with our reference κ = −0.2, Fig. 4b. While in our reference system bundles behave like contractile objects tending to straighten, for κ = −0.8 they behave like extensile objects enhancing curvature (see blue insets), which results in continuous defect nucleation as highly bent bundles destabilize and fragment, as well as defect annihilation as pairs of free ends merge to reorganize the network (see second inset in (b,II)), in a behavior akin to active turbulence [85]. See Movie 5 for an illustration and for the slightly extensile case κ = −0.5. To further substantiate this contractile vs extensile behavior of the nematic structures emerging from a contractile active gel with extensile deviatoric behavior, we plotted the difference between the stress along the nematic direction and perpendicular to it, σ|| − σ⊥. We note that for κ < 0 as required to obtain fibrillar patterns, the active component of this quantity is negative, σ||act−σ⊥act<0 , Fig. 1c. In the case of bundles with contractile phenomenology, i.e. a tendency to straighten (b,I), we found that σ|| − σ⊥ was positive on the dense bundles and nearly zero in between, indicating that bundles are contractile structures embedded in a largely isotropic matrix. Conversely, in bundles with extensile phenomenology (b,II), we found intricate stress patterns with regions of negative σ|| − σ⊥. The extensile tension pattern for κ = −0.8 is more clearly appreciated in 1D simulations, Supp. Fig. 2, where the dense nematic structures cannot fragment as a result of the bend-type instability. This figure also shows how the competition between active and viscous stresses dictates the emergent contractile vs extensile behavior of the nematic bundles. In the contractile case (κ = −0.2), even if σ||act<σ⊥act , the negative viscous stresses due to the self-reinforcing flows are significantly larger in magnitude perpendicular to the nematic direction, resulting in σ|| − σ⊥. In summary, our theory predicts that a contractile nematic active gel can self-organize into fibrillar patterns with mesoscale bundles that are either contractile or extensile depending on the parameter regime. This complex interplay between the contractility vs extensibility of the base material and that of the mesoscale nematic pattern resonates with recent observations where tissues made of contractile cells behave collectively as extensile nematic systems [45, 86].

Focusing on contractile bundles, we then examined a different parameter regime known to trigger sustained pattern dynamics. For isotropic gels, previous work has shown that reducing friction triggers chaotic dynamics as the distance between high-density regions, ℓpattern = 2π/vcrit, becomes comparable or smaller than the hydrodynamic length scale [29], thus enabling their mechanical interaction. In a model devoid of orientational order, reducing friction is equivalent to increasing k¯d . In our model, however, we can either reduce γ, which in non-dimensional terms means increasing k¯d,a¯,b¯ and λ¯⊙ . in concert, or increase k¯d while leaving all other non-dimensional parameters fixed. The first of these choices leads to dynamic and hierarchical networks with very dense and highly-aligned bundles, which coexist with perpendicular weak bundles with much lower density enrichment and ordering. These two families of bundles enclose cells of isotropic and sparse gel, Fig. 4c(I). Junctions where two or more dense bundles meet are very unfavorable and short-lived, but junctions of two dense and a weak bundle are much more stable. Because bundles are mechanically coupled, the network actively reorganizes in events where dense bundles become weak bundles and vice-versa, inset and Movie 6(I). We note that here σ|| > σ⊥, Fig. 4c(I) and Supp. Fig. 2d, and hence the persistent dynamics are unrelated to the previously described behavior of extensile bundles.

The second choice to favor mechanical interaction of bundles, increasing k¯d only, leads to very different networks with high-density aster-like clusters interconnected by straight actin bundles. Because now λ¯⊙ is not particularly large, order is low at the core of these clusters, enabling high-valence networks where four bundles often meet at one cluster. For k¯d = 10, the network is stable and nearly crystalline, Fig. 4c(II), whereas for k¯d = 20, it becomes highly dynamical and pulsatile with frequent collapse of polygonal cells by fusion of neighboring actin clusters and their attached bundles (black polygons) and nucleation of new bundles within large low-density cells (dashed/solid green lines), Fig. 4c(III) and Movie 6(III). This architecture and dynamics resemble those of early C. elegans embryos [87], adherent epithelial cells treated with epidermal growth factor [27] and mouse embryonic stem cells [83]. Recent active gel models accounting for RhoA signaling develop similar pulsatile behaviors in 2D, but do not predict the orientational order of the spatio-temporal patterns of the actomyosin cortex [88].

In summary, our theory maps how effective parameters of the actin gel control the active self-organization of a uniform and isotropic gel into a pattern of high-density nematic bundles embedded in a low-density isotropic matrix, including the activity threshold, the bundle spacing, orientation, connectivity and dynamics.

Assessment of activity parameters κ and λ⊙ through discrete network simulations.

(a) Illustration of the computational domain of the discrete network as a uniform representative volume element of the gel. (b) Sketch of model ingredients and setup to compute tension along and perpendicular to the nematic direction. (c) Typical time-signal for parallel and perpendicular tensions following addition of crosslinkers and motors (translucent lines) along with time average (solid lines) for isotropic and anisotropic networks. Tension is normalized by mean tension σ¯=(σ||+σ⊥)/2 computed from time-averages and time by actin turnover time. (d) Mean tension as a function of network density for several nematic parameters S0, where both quantities are normalized by their values for the lowest density. With this normalization, Eq. (11) predicts a linear dependence with slope λ = 1 (dashed line). Error bars span two standard deviations. (e) Deviatoric tension as a function of nematic order for different densities. The dashed line is a linear regression to simulation data. (f) Dynamics of nematic order in a periodic network following addition of crosslinkers and motors for three initial values of nematic order. (g) Rate of change of nematic order normalized by turnover rate as a function of initial nematic order at zero and finite temperature.

Microscopic origin of κ < 0 and λ⊙ > 0 through discrete network simulations

A central prediction of our model is that the self-organization of nematic bundles, the most prominent emerging organization in actin gels across cell types and length-scales, requires that active tension perpendicular to nematic orientation is larger than along this direction (κ < 0, see Fig. 1c), at least at the onset of pattern formation. This fact may seem counter-intuitive in that dense nematic bundles are associated with large contractile tension along their axis, but as discussed in Fig. 4c(I), even if κ < 0 bundles can be contractile because of the interplay between active and viscous stresses. Another prediction of our model is that the active assembly dense nematic bundles requires active alignment controlled by parameter λ⊙ > 0, Fig. 3. Being central to our conclusions, we then sought to examine the plausibility of the conditions κ < 0 and λ⊙ > 0 of our phenomenological theory by performing discrete network simulations using the open-source cytoskeleton simulation suite Cytosim [89].

Similarly to reconstituted actomyosin gels, discrete network simulations of the actin cytoskeleton tend to irreversibly collapse into clumps [70], although proper tuning of model parameters can lead to sustained contractile states [90]. However, to our knowledge, these models have not been able to reproduce the heterogeneous nonequilibrium contractile states involving sustained self-reinforcing flows underlying the pattern formation mechanism studied here. For this reason, rather than trying to reproduce the assembly and maintenance of patterns of dense nematic structures, we specifically focused on investigating whether the average behavior of the discrete network is compatible with the conditions κ < 0 and λ⊙ > 0 of our continuum theory. For this, we studied the buildup of tension and dynamics of nematic order in uniform representative volume elements of varying density and filament alignment after being brought out of equilibrium, Fig. 5a. Each of our simulated volume elements can represent a uniform system prior to pattern formation, or a material point in the actomyosin gel.

We performed 2D simulations in which semi-flexible filaments interact with crosslinkers and myosin motors, all of which undergo turnover and have a stoichiometry previously used to model the actin cytoskeleton [91], Fig. 5b. See Appendix D for a detailed description of the simulation protocol. Briefly, we modified Cytosim to account for average orientational order in the simulation box, which we evaluated as a sample average of orientations over the ensemble of segments composing the filaments. We further introduced a nematic energy penalty in the network allowing us to restrain average nematic order to a target value S0.

We first prepared a system consisting only of randomly oriented actin fibers, imposed the desired orientational order S0 using the nematic penalty and equilibrated the system. In a first set of simulations, once S0 was reached, we deactivated the nematic penalty and added crosslinkers and myosins, driving the system out-of-equilibrium. The free contraction of the system was prevented by the addition of anchors at the boundary of the representative volume element, which also allowed us to compute anchor forces and hence estimate the effective active tension along the nematic direction σ|| = σijσiσj and perpendicular to it, σ⊥ = σijσimj with nimi = 0 and mimi = 1, Fig. 5b.

Addition of crosslinkers and myosins leads to bundling of actin filaments at the microscale, Movie 7, distinct from the mesoscale fibrillar pattern formation emerging from the active gel model. It also leads to out-of-equilibrium tension as measured by the anchors. For isotropic networks (S0 = 0), active tension is isotropic with σ|| ≈ σ⊥. For anisotropic networks, however, we found that tension becomes anisotropic with σ⊥ > σ||, Fig. 5c.

We systematically characterized this behavior varying initial orientational order and network density. According to our active gel model, Eqs. (5,6,7), in the absence of nematic gradients and flow, the tension components σ|| and σ⊥ satisfy the following relations in terms of mean and deviatoric tensions σ¯=(σ∥+σ⊥)/2=λρ and (σ∥−σ⊥)/σ¯=κS. (11)

Remarkably, our discrete network simulations closely followed these relations, Fig. 5d,e, which allowed us to estimate κ ≈ −1.6. We robustly found that κ < 0 for perturbations of selected parameters of the discrete network model as long as turnover rates of crosslinkers and myosins were relatively fast.

We then wondered if the discrete network simulations could provide evidence for the orientational activity parameter in our theory, ρλ⊙. For a uniform system with nematic order along a given direction and ignoring the susceptibility parameter b, Eq. (8) becomes ηrotS˙+(2a−ρλ⊙)S+κS(S−S0)=0, (12)

where ηrot is the viscous drag of the filaments in the discrete network simulations, a > 0 the entropic tendency of the model to return to isotropy, ρλ⊙ the active forcing of nematic order resulting from crosslinkers and motors, and the last term accounts for the effect of the nematic penalty with coefficient KS. As a first test of this model, we started from an isotropic and periodic network and tracked the athermal dynamics of S under the action of the nematic penalty in the absence of anchors, crosslinkers and myosins. For λ⊙ = 0 and a = 0, Eq. (12) predicts an exponential relaxation given by S(t) = S0(1 − e−KSt/ηrot), which very closely matched the simulation data for different values of So and for a single fitting parameter nrot, Supp. Fig.3(III) and Movie 8. We then deactivated the nematic penalty and added crosslinkers and motors, but not anchors, to track unconstrained dynamics of nematic order starting from different values of S0. In agreement with the notion of an active force driving nematic order, we found that S(t) monotonically increased, Fig. 5f. More quantitatively, we tested the short-time prediction of Eq. (12), ηrotS˙=(ρλ⊙−2a)S0 , by plotting S˙ as estimated from our simulations, as a function of S0, Fig. 5g. We found a nearly linear relation with positive slope, hence providing evidence for an active generalized force driving order. In agreement with the theory, in the athermal limit, the tendency to actively increase order is faster as the entropic tendency to disorder is absent (a = 0).

In summary, discrete network cytoskeletal simulations provide a microscopic justification for two key ingredients of our active gel theory, namely that nematic order elicits (1) anisotropic active tensions, which can be larger perpendicular to the nematic direction (κ < 0), and (2) active generalized forces driving further ordering..

Conclusions

We have developed a theory for the active self-organization of initially uniform and isotropic actin gels into various localized dense nematic architectures embedded in an isotropic matrix of low density. This model predicts a variety of emergent patterns involving asters, tactoids, and bands with perpendicular nematic organization. More importantly, it identifies a wide parameter space where the active gel spontaneously develops patterns of dense nematic bundles, the most prominent nematic architecture across scales and cell types. Contrary to previous works on instabilities and pattern formation in active nematics, the mechanism proposed here relies on the advective instability and compressible self-reinforcing flows typical of actomyosin gels. We have characterized how the activity threshold, spacing, geometry, connectivity and dynamics these patterns depends on effective active gel parameters. Because these mesoscale parameters depend on the composition and dynamics of the network at the molecular scale, our results portray actin gels as responsive and reconfigurable active materials with an intrinsic ability to assemble patterns of nematic bundles that cells can finely regulate.

We have further shown that the spontaneous tendency of the gel to assemble bundle patterns can be directed via subtle cues. Thus, a combination of biochemical control of actin dynamics along with geometric, mechanical or biochemical guiding [92, 93] may explain the emergence and context-dependent organization of regular patterns of bundle networks, from sub-cellular to organism scales [12, 14, 19, 21, 22, 25–28, 30]. Consistently, perturbations of myosin contractility have been shown to alter, disorder, or even prevent the formation patterns of parallel bundles in C. elegans [19], whereas perturbations of actin polymerization in drosophila embryos impair the robust organization of actin bundle patterns at the cellular and organ scales by disrupting orientation and spacing, but not the intrinsic tendency of the actomyosin cytoskeleton to form patterns of parallel bundles [21]. In adherent cells, the patterns of dense nematic bundles presented here may act as precursors of stress fibers as the actomyosin cytoskeleton interacts with the focal adhesion machinery in a cellular domain with boundary conditions set by the polymerization velocity at the leading edge and the nucleus.

Our theory identifies two key requirements on activity parameters for the self-organization of patterns of nematic bundles, namely active tension anisotropy with larger tension perpendicular to the nematic direction and generalized active forces tending to increase nematic order. Complementarily to our phenomenological theory, we have examined the plausibility of these conditions with discrete network simulations of homogeneous representative volume elements of different density and orientational order, which have confirmed our constitutive assumptions. We expect, however, that in a different parameter regime anisotropic tension may be larger along the nematic direction (κ > 0). For instance, once bundles are dense and maximally aligned, the ability of the active nematic gel to perform active tension perpendicular to the nematic direction may saturate, while myosin motors may contract the gel along the nematic direction more effectively. The regime studied here explains the initial assembly of dense nematic bundles, but not their maturation to become highly contractile or viscoelastic as demonstrated for stress fibers depending on different isoforms of nonmuscle myosin II or on actin regulators such as zyxin [94, 95]. Our work thus suggests further experimental and computational work to establish a comprehensive mapping between molecular and mesoscale properties of the active gel, and how these properties control the emergent network architecture and mechanical properties.